---
title: "Clam resazurin metabolic rate analysis – 40°C Control vs Selected"
author: "SR Roberts"
date: '2026-03-30'
output:
  github_document: null
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 6
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(cowplot)
library(tools)
```

# Load & parse data

Each file is a BioTek Synergy HTX export for a 24-well plate.
Columns 1-5 contain individual clam samples; column 6 is blank (resazurin
working solution, no animal). Two plates per treatment (blue / red).

```{r parse-function}
parse_plate_24well <- function(file) {
  lines <- readLines(file, warn = FALSE)
  results_idx <- which(trimws(lines) == "Results")
  if (length(results_idx) == 0) stop("No 'Results' in ", file)

  idx <- results_idx[1]
  j <- idx + 2
  data_lines <- character()
  while (j <= length(lines)) {
    line <- lines[j]
    if (nchar(trimws(line)) == 0) break
    if (!grepl("^[A-Da-d]\\t", line)) break
    data_lines <- c(data_lines, line)
    j <- j + 1
  }

  map_dfr(data_lines, function(row) {
    parts  <- unlist(strsplit(row, "\t"))
    row_id <- trimws(parts[1])
    vals   <- suppressWarnings(as.numeric(parts[2:(length(parts) - 1)]))
    tibble(
      row_letter = row_id,
      col        = seq_along(vals),
      fluor      = vals
    )
  })
}
```

```{r load-files}
data_dir <- file.path(
  rprojroot::find_rstudio_root_file(),
  "data", "clam", "20260330-clam-40C-C-vs-S"
)
files <- list.files(data_dir, pattern = "\\.txt$", full.names = TRUE)

all_data <- map_dfr(files, function(f) {
  fn <- tolower(basename(f))
  treatment <- if (grepl("plate-c", fn)) "Control" else "Selected"
  plate_col <- str_extract(fn, "(?<=plate-[cs]-)(blue|red)")
  tp        <- as.numeric(str_extract(fn, "(?<=-t)\\d+\\.\\d+"))
  parse_plate_24well(f) %>%
    mutate(treatment = treatment, plate_color = plate_col, timepoint = tp,
           well = paste0(row_letter, col))
})

all_data <- all_data %>%
  mutate(
    well_type = if_else(col == 6, "blank", "sample"),
    plate_id  = paste(treatment, plate_color, sep = "_")
  )

head(all_data)
```

# Data quality check

```{r qc-blanks}
blanks <- all_data %>% filter(well_type == "blank")

ggplot(blanks, aes(x = factor(timepoint), y = fluor, colour = plate_id)) +
  geom_jitter(width = 0.15, size = 2) +
  labs(x = "Timepoint (h)", y = "Blank fluorescence (RFU)",
       colour = "Plate", title = "Blank wells (column 6) over time") +
  theme_classic()
```

Flag: Some S-red blank wells (rows B, C) show elevated fluorescence at later
time points (potential contamination / wicking from adjacent sample wells).

```{r qc-duplicate}
s_blue_t0 <- all_data %>%
  filter(treatment == "Selected", plate_color == "blue", timepoint == 0) %>%
  select(row_letter, col, fluor)
s_red_t0 <- all_data %>%
  filter(treatment == "Selected", plate_color == "red", timepoint == 0) %>%
  select(row_letter, col, fluor)

if (identical(s_blue_t0, s_red_t0)) {
  message("NOTE: plate-S-red-T0.0 and plate-S-blue-T0.0 contain identical data ",
          "(same plate number and timestamp). The S-red T0 data may be a copy error.")
}
```

# Blank correction

Subtract per-plate, per-timepoint mean blank from every sample well.

```{r blank-correct}
blank_means <- all_data %>%
  filter(well_type == "blank") %>%
  group_by(plate_id, timepoint) %>%
  summarise(blank_mean = mean(fluor, na.rm = TRUE), .groups = "drop")

samples <- all_data %>%
  filter(well_type == "sample") %>%
  left_join(blank_means, by = c("plate_id", "timepoint")) %>%
  mutate(fluor_corr = fluor - blank_mean)

head(samples)
```

# Summary statistics

```{r summary-stats}
plate_summary <- samples %>%
  group_by(treatment, plate_color, plate_id, timepoint) %>%
  summarise(
    n        = n(),
    mean_rfu = mean(fluor_corr, na.rm = TRUE),
    sd_rfu   = sd(fluor_corr,   na.rm = TRUE),
    se_rfu   = sd_rfu / sqrt(n),
    .groups  = "drop"
  )

treatment_summary <- samples %>%
  group_by(treatment, timepoint) %>%
  summarise(
    n        = n(),
    mean_rfu = mean(fluor_corr, na.rm = TRUE),
    sd_rfu   = sd(fluor_corr,   na.rm = TRUE),
    se_rfu   = sd_rfu / sqrt(n),
    .groups  = "drop"
  )

treatment_summary
```

# Plots

```{r fig-dir}
fig_dir <- file.path(
  rprojroot::find_rstudio_root_file(),
  "figures", "clam", "20260330-clam-40C-C-vs-S"
)
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
```

## Individual well trajectories by plate

```{r plot-individual, fig.width=10, fig.height=6}
p_individual <- samples %>%
  ggplot(aes(x = timepoint, y = fluor_corr,
             colour = well, group = well)) +
  facet_wrap(~plate_id) +
  geom_point(size = 1) +
  geom_line(alpha = 0.6) +
  labs(x = "Time at 40°C (h)", y = "Blank-corrected fluorescence (RFU)") +
  theme_classic() +
  theme(legend.position = "none")

p_individual

ggsave(file.path(fig_dir, "individual_wells.png"),
       p_individual, width = 10, height = 6)
```

## Plate-level means

```{r plot-plate-means, fig.width=7, fig.height=5}
p_plate <- plate_summary %>%
  ggplot(aes(x = timepoint, y = mean_rfu,
             colour = plate_id, fill = plate_id)) +
  geom_ribbon(aes(ymin = mean_rfu - se_rfu,
                  ymax = mean_rfu + se_rfu), alpha = 0.2, colour = NA) +
  geom_point() +
  geom_line() +
  labs(x = "Time at 40°C (h)", y = "Mean blank-corrected RFU ± SE",
       colour = "Plate", fill = "Plate") +
  theme_classic()

p_plate

ggsave(file.path(fig_dir, "plate_means.png"), p_plate, width = 7, height = 5)
```

## Treatment-level comparison: Control vs Selected

```{r plot-treatment, fig.width=7, fig.height=5}
p_treatment <- treatment_summary %>%
  ggplot(aes(x = timepoint, y = mean_rfu,
             colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = mean_rfu - se_rfu,
                  ymax = mean_rfu + se_rfu), alpha = 0.2, colour = NA) +
  geom_point(size = 2.5) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("Control" = "#377eb8", "Selected" = "#e41a1c")) +
  scale_fill_manual(values   = c("Control" = "#377eb8", "Selected" = "#e41a1c")) +
  labs(x = "Time at 40°C (h)",
       y = "Mean blank-corrected fluorescence (RFU ± SE)",
       colour = "Treatment", fill = "Treatment",
       title = "Resazurin metabolic activity: Control vs Selected clams at 40°C") +
  theme_classic(base_size = 13)

p_treatment

ggsave(file.path(fig_dir, "treatment_CvsS.png"), p_treatment, width = 7, height = 5)
```

## Rate of fluorescence increase (metabolic rate proxy)

Compute the slope of fluorescence over time for each well (linear fit across
all time points) as a simple metabolic rate metric.

```{r metabolic-rate, fig.width=6, fig.height=5}
rates <- samples %>%
  group_by(treatment, plate_color, plate_id, well) %>%
  summarise(
    slope = coef(lm(fluor_corr ~ timepoint))[2],
    .groups = "drop"
  )

p_rate <- rates %>%
  ggplot(aes(x = treatment, y = slope, fill = treatment)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.4, width = 0.5) +
  geom_jitter(aes(colour = plate_color), width = 0.15, size = 1.5, alpha = 0.7) +
  scale_fill_manual(values   = c("Control" = "#377eb8", "Selected" = "#e41a1c")) +
  scale_colour_manual(values = c("blue" = "steelblue", "red" = "tomato")) +
  labs(x = NULL, y = "Fluorescence rate of change (RFU / h)",
       fill = "Treatment", colour = "Plate",
       title = "Metabolic rate: Control vs Selected") +
  theme_classic(base_size = 13)

p_rate

ggsave(file.path(fig_dir, "metabolic_rate_boxplot.png"),
       p_rate, width = 6, height = 5)
```

```{r rate-stats}
rate_summary <- rates %>%
  group_by(treatment) %>%
  summarise(
    n          = n(),
    mean_slope = mean(slope),
    sd_slope   = sd(slope),
    se_slope   = sd_slope / sqrt(n),
    .groups    = "drop"
  )

rate_summary

t_test <- t.test(slope ~ treatment, data = rates)
t_test
```
