---
title: "Clam resazurin metabolic rate analysis"
author: "AS Huffmyer"
date: '2026'
output:
  github_document: null
  md_document: default
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 6
    toc_float: true
editor_options: 
  chunk_output_type: console
---

This script analyzes resazurin assays for clam samples using a 48-well plate format (BioTek Synergy HTX reader). Column 1 wells contain clam samples; column 2 wells contain blanks.

# Set up 

Set up workspace, set options, and load required packages.    
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries. 
```{r}
library(tidyverse)
library(ggplot2)
library(cowplot)
library(tools)
```

# Load data 

## Parse plate reader files

Define a function to parse BioTek Synergy HTX .txt files. Each file may contain one or more plate readings separated by a "Results" header. For files with multiple readings the fluorescence values are averaged across readings.

```{r}
parse_plate_file <- function(file) {
  lines <- readLines(file, warn = FALSE)
  
  # Find all lines that contain only "Results"
  results_idx <- which(trimws(lines) == "Results")
  
  if (length(results_idx) == 0) {
    stop(paste("No 'Results' section found in file:", file))
  }
  
  readings_list <- list()
  
  for (idx in results_idx) {
    # Line idx   = "Results"
    # Line idx+1 = column header (	1	2	3	4	5	6	7	8)
    # Lines idx+2 to idx+7 = data rows A-F
    data_lines <- lines[(idx + 2):(idx + 7)]
    
    plate_data <- lapply(data_lines, function(row) {
      parts <- unlist(strsplit(row, "\t"))
      row_letter <- trimws(parts[1])
      # parts[2:9] are the 8 well values; the last element is the filter label
      values <- suppressWarnings(as.numeric(parts[2:9]))
      data.frame(
        Row            = row_letter,
        sample_fluor   = values[1],  # column 1 = clam sample well
        blank_fluor    = values[2],  # column 2 = blank well
        stringsAsFactors = FALSE
      )
    })
    
    readings_list[[length(readings_list) + 1]] <- bind_rows(plate_data)
  }
  
  # If multiple readings exist, average col1 and col2 per row
  if (length(readings_list) > 1) {
    bind_rows(readings_list, .id = "reading") %>%
      group_by(Row) %>%
      summarise(
        sample_fluor = mean(sample_fluor, na.rm = TRUE),
        blank_fluor  = mean(blank_fluor,  na.rm = TRUE),
        .groups = "drop"
      )
  } else {
    readings_list[[1]]
  }
}
```

Read and combine all plate files from `data/clam/`.

```{r}
folder_path <- "data/clam"

file_list <- list.files(path = folder_path, pattern = "\\.txt$",
                        full.names = TRUE, ignore.case = TRUE)

if (length(file_list) == 0) {
  stop("No .txt files found in data/clam. Check folder path.")
}

data_list <- list()

for (file in file_list) {
  file_name <- file_path_sans_ext(basename(file))
  
  # Extract plate id (plate1, plate2) and timepoint (0, 1, 3, 4) from filename
  plate     <- str_extract(tolower(file_name), "plate\\d+")
  timepoint <- str_extract(tolower(file_name), "t\\d+") %>%
    str_remove("t") %>%
    as.numeric()
  
  parsed <- parse_plate_file(file) %>%
    mutate(
      plate     = plate,
      timepoint = timepoint,
      well      = paste0(Row, "1")
    )
  
  data_list[[file_name]] <- parsed
}

combined_data <- bind_rows(data_list)

head(combined_data)
```

## Normalize to T0

Divide each well's fluorescence by the T0 (baseline) value to express fluorescence as fold-change relative to the start of the assay.

```{r}
combined_data <- combined_data %>%
  group_by(plate, well) %>%
  arrange(timepoint, .by_group = TRUE) %>%
  mutate(
    sample_norm = sample_fluor / first(sample_fluor),
    blank_norm  = blank_fluor  / first(blank_fluor)
  ) %>%
  ungroup()

head(combined_data)
```

## Blank correction

Calculate the mean normalized blank fluorescence per plate and timepoint, then subtract it from the normalized sample fluorescence (adding 1 to keep values centred at 1 at T0).

```{r}
blanks <- combined_data %>%
  group_by(plate, timepoint) %>%
  summarise(mean_blank_norm = mean(blank_norm, na.rm = TRUE), .groups = "drop")

combined_data <- left_join(combined_data, blanks, by = c("plate", "timepoint")) %>%
  mutate(fluorescence_corr = sample_norm - mean_blank_norm + 1)

head(combined_data)
```

# Plotting

## Raw fluorescence over time

```{r}
plot_raw <- combined_data %>%
  ggplot(aes(x = timepoint, y = sample_fluor,
             colour = plate, group = interaction(plate, well))) +
  geom_point() +
  geom_line() +
  scale_colour_manual(values = c("plate1" = "steelblue", "plate2" = "darkorange")) +
  labs(
    x      = "Timepoint (h)",
    y      = "Raw Fluorescence",
    colour = "Plate"
  ) +
  theme_classic()

plot_raw

ggsave(plot_raw, filename = "figures/clam/raw_fluorescence.png", width = 6, height = 4)
```

## Normalized fluorescence over time

```{r}
plot_norm <- combined_data %>%
  ggplot(aes(x = timepoint, y = sample_norm,
             colour = plate, group = interaction(plate, well))) +
  geom_point() +
  geom_line() +
  scale_colour_manual(values = c("plate1" = "steelblue", "plate2" = "darkorange")) +
  labs(
    x      = "Timepoint (h)",
    y      = "Fluorescence (fold change from T0)",
    colour = "Plate"
  ) +
  theme_classic()

plot_norm

ggsave(plot_norm, filename = "figures/clam/normalized_fluorescence.png", width = 6, height = 4)
```

## Blank-corrected fluorescence over time

```{r}
plot_corr <- combined_data %>%
  ggplot(aes(x = timepoint, y = fluorescence_corr,
             colour = plate, group = interaction(plate, well))) +
  geom_point() +
  geom_line() +
  scale_colour_manual(values = c("plate1" = "steelblue", "plate2" = "darkorange")) +
  labs(
    x      = "Timepoint (h)",
    y      = "Blank-corrected Fluorescence",
    colour = "Plate"
  ) +
  theme_classic()

plot_corr

ggsave(plot_corr, filename = "figures/clam/corrected_fluorescence.png", width = 6, height = 4)
```

## Mean metabolic activity over time

Plot group means ± SE.

```{r}
summary_data <- combined_data %>%
  group_by(plate, timepoint) %>%
  summarise(
    mean_fluor = mean(fluorescence_corr, na.rm = TRUE),
    se_fluor   = sd(fluorescence_corr,   na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

plot_mean <- summary_data %>%
  ggplot(aes(x = timepoint, y = mean_fluor,
             colour = plate, fill = plate)) +
  geom_ribbon(aes(ymin = mean_fluor - se_fluor, ymax = mean_fluor + se_fluor),
              alpha = 0.2, colour = NA) +
  geom_point() +
  geom_line() +
  scale_colour_manual(values = c("plate1" = "steelblue", "plate2" = "darkorange")) +
  scale_fill_manual(values   = c("plate1" = "steelblue", "plate2" = "darkorange")) +
  labs(
    x      = "Timepoint (h)",
    y      = "Blank-corrected Fluorescence (mean ± SE)",
    colour = "Plate",
    fill   = "Plate"
  ) +
  theme_classic()

plot_mean

ggsave(plot_mean, filename = "figures/clam/mean_fluorescence.png", width = 6, height = 4)
```

## Individual well trajectories faceted by plate

```{r}
plot_facet <- combined_data %>%
  ggplot(aes(x = timepoint, y = fluorescence_corr,
             colour = well, group = well)) +
  facet_wrap(~plate) +
  geom_point() +
  geom_line() +
  labs(
    x      = "Timepoint (h)",
    y      = "Blank-corrected Fluorescence",
    colour = "Well"
  ) +
  theme_classic()

plot_facet

ggsave(plot_facet, filename = "figures/clam/individual_wells.png", width = 8, height = 4)
```
