---
title: "Clam Dashboard"
subtitle: "Auto-updated from plate exports in data/clam"
format: html
resources:
  - output/clam/dashboard-data.csv
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(dplyr)
library(readr)
library(ggplot2)
library(knitr)

data_path <- file.path("output", "clam", "dashboard-data.csv")
qc_path <- file.path("output", "clam", "dashboard-qc.csv")

if (!file.exists(data_path)) {
  cat("## No dashboard data yet\n\n")
  cat("Run scripts/clam/build_clam_dashboard_data.R to generate data.\n")
  knitr::knit_exit()
}

dat <- readr::read_csv(data_path, show_col_types = FALSE)
qc <- if (file.exists(qc_path)) readr::read_csv(qc_path, show_col_types = FALSE) else tibble()

if (nrow(dat) == 0) {
  cat("## Dashboard dataset is empty\n\n")
  cat("No matching plate files were discovered under data/clam.\n")
  knitr::knit_exit()
}

dat <- dat %>%
  mutate(
    experiment_plate = paste(experiment_dir, plate_id, sep = " | "),
    sample_group = dplyr::coalesce(sample_label, well_id)
  )
```

## What This Shows

- Plate-wide trajectories across time for every well.
- Per-well trajectories for all wells (filterable by experiment in the static report by scrolling to the relevant facets).
- QC table with parse status and warning flags.

## Interactive Explorer

Use these selectors to focus on one experiment, one plate, or specific wells.

```{ojs}
dashboard = await FileAttachment("output/clam/dashboard-data.csv").csv({typed: true})

experimentValues = [...new Set(dashboard.map(d => d.experiment_dir))]

experimentSortKey = (name) => {
  const m = /^([0-9]{8})/.exec(name)
  if (!m) return Number.NEGATIVE_INFINITY
  return Number.parseInt(m[1], 10)
}

mostRecentExperiment = experimentValues
  .slice()
  .sort((a, b) => {
    const ka = experimentSortKey(a)
    const kb = experimentSortKey(b)
    if (ka !== kb) return kb - ka
    return a.localeCompare(b)
  })[0]

experiments = ["All", ...experimentValues]
viewof experimentSel = Inputs.select(experiments, {
  label: "Experiment",
  value: mostRecentExperiment || "All"
})

plateCandidates = dashboard
  .filter(d => experimentSel === "All" || d.experiment_dir === experimentSel)
plates = ["All", ...new Set(plateCandidates.map(d => d.plate_id))]
viewof plateSel = Inputs.select(plates, {
  label: "Plate",
  value: "All"
})

wellCandidates = dashboard
  .filter(d => (experimentSel === "All" || d.experiment_dir === experimentSel) &&
               (plateSel === "All" || d.plate_id === plateSel))
wells = [...new Set(wellCandidates.map(d => d.well_id))].sort()
defaultWells = wells.slice(0, Math.min(6, wells.length))

viewof wellSel = Inputs.select(wells, {
  label: "Well(s)",
  multiple: true,
  value: defaultWells
})

filtered = dashboard.filter(d =>
  (experimentSel === "All" || d.experiment_dir === experimentSel) &&
  (plateSel === "All" || d.plate_id === plateSel) &&
  (wellSel.length === 0 || wellSel.includes(d.well_id))
)

Plot.plot({
  height: 420,
  grid: true,
  x: {label: "Time (hours)"},
  y: {label: "Fluorescence"},
  color: {legend: true, label: "Well"},
  marks: [
    Plot.line(filtered, {
      x: "time_hr",
      y: "value",
      stroke: "well_id",
      z: d => `${d.experiment_dir}|${d.plate_id}|${d.well_id}`
    }),
    Plot.dot(filtered, {
      x: "time_hr",
      y: "value",
      fill: "well_id",
      r: 2.5,
      tip: true
    })
  ]
})
```

## Plate-Wide View

```{r}
#| echo: false

p_plate <- ggplot(dat, aes(x = time_hr, y = value, color = well_id, group = well_id)) +
  geom_line(alpha = 0.7) +
  geom_point(size = 1.5, alpha = 0.8) +
  facet_wrap(~experiment_plate, scales = "free_y") +
  labs(
    x = "Time (hours)",
    y = "Fluorescence",
    color = "Well"
  ) +
  theme_bw(base_size = 11)

print(p_plate)
```

## Individual Well View

```{r}
#| echo: false

p_well <- ggplot(dat, aes(x = time_hr, y = value, group = interaction(experiment_plate, well_id))) +
  geom_line(color = "#1f78b4", alpha = 0.4) +
  geom_point(color = "#1f78b4", alpha = 0.6, size = 1.2) +
  facet_wrap(~well_id, scales = "free_y") +
  labs(
    x = "Time (hours)",
    y = "Fluorescence",
    title = "Trajectories by Well Across Plates and Experiments"
  ) +
  theme_bw(base_size = 11)

print(p_well)
```

## Latest Snapshot by Well

```{r}
#| echo: false

latest <- dat %>%
  group_by(experiment_dir, plate_id, well_id) %>%
  filter(time_hr == max(time_hr, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(experiment_dir, plate_id, well_id) %>%
  select(experiment_dir, plate_id, well_id, time_hr, value, sample_label, treatment, layout_status)

knitr::kable(latest, caption = "Latest fluorescence value per well")
```

## Parse And Data Quality Status

```{r}
#| echo: false

if (nrow(qc) > 0) {
  qc_show <- qc %>%
    arrange(desc(parse_ok), experiment_dir, source_file)
  knitr::kable(qc_show, caption = "File-level parse status")
} else {
  cat("No QC data available.\n")
}
```
