---
title: "01.00-resazurin-20260528-volume_testing"
author: "Sam White"
date: "2026-05-28"
output:
  github_document:
    toc: true
    number_sections: true
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

# BACKGROUND

Comparison of how different resazurin solution volumes affect fluorescence
readings. All wells contain the same resazurin working solution; only the volume loaded per well differs. Volumes tested: 100,
200, 300, 400, 500, 750, 900, and 1000 µL. Six replicate wells per volume.
Fluorescence was read at T0 on a Synergy HTX (Agilent) plate reader.


## Expected inputs

| Path | Description |
|------|-------------|
| `data/volume_testing/20280528/plate-A-T*.txt` | Plate reader fluorescence exports |
| `data/volume_testing/20280528/layout.csv` | Well metadata: plate ID, well, volume group |

## Expected outputs

All outputs are written to `output/volume_testing/20260528/`.

| File | Description |
|------|-------------|
| `figures/` | All plots generated by this script |
| `volume_fluorescence.csv` | Per-well raw fluorescence with volume metadata |
| `volume_summary.csv` | Volume-level summary statistics |
| `pairwise_stats.csv` | Tukey-adjusted pairwise comparisons |

# SETUP

## Knitr

```{r setup}
knitr::opts_chunk$set(
  echo    = TRUE,
  eval    = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "",
  results = 'hold'
)
```

## Libraries
```{r libraries}
library(tidyverse)
library(emmeans)
library(multcompView)
```

## Helper Functions

```{r helper-fns}
normalize_well_id <- function(x) {
  x <- toupper(trimws(x))
  valid <- str_detect(x, "^[A-Z]+[0-9]+$")
  out <- rep(NA_character_, length(x))
  if (!any(valid)) return(out)
  m <- str_match(x[valid], "^([A-Z]+)([0-9]+)$")
  out[valid] <- paste0(m[, 2], as.integer(m[, 3]))
  out
}

parse_time_hr <- function(path) {
  hit <- str_match(basename(path),
                   "(?i)-T([0-9]+(?:\\.[0-9]+)?)\\.txt$")
  as.numeric(hit[, 2])
}

parse_plate_id <- function(path) {
  hit <- str_match(basename(path),
    "(?i)^plate-([A-Za-z0-9-]+)-T[0-9]+(?:\\.[0-9]+)?\\.txt$")
  id <- hit[, 2]
  ifelse(is.na(id), "unknown", id)
}

extract_results_block <- function(lines) {
  results_idx <- which(trimws(lines) == "Results")
  if (length(results_idx) == 0) stop("No Results section found")
  idx <- results_idx[1]
  header_tokens <- str_split(lines[idx + 1], "\\t")[[1]] |> trimws()
  col_ids <- header_tokens[
    header_tokens != "" & str_detect(header_tokens, "^[0-9]+$")]
  j <- idx + 2
  data_lines <- character()
  while (j <= length(lines)) {
    line <- lines[j]
    if (trimws(line) == "") break
    if (!str_detect(line, "^[A-Za-z]\\t")) break
    data_lines <- c(data_lines, line)
    j <- j + 1
  }
  list(col_ids = col_ids, data_lines = data_lines)
}

parse_plate_export <- function(path) {
  lines <- readLines(path, warn = FALSE)
  res   <- extract_results_block(lines)

  map_dfr(res$data_lines, function(line) {
    tokens <- str_split(line, "\\t")[[1]] |> trimws()
    tokens <- tokens[tokens != ""]
    row_letter <- tokens[1]
    nums <- suppressWarnings(as.numeric(tokens[-1]))
    valid_idx <- which(!is.na(nums))
    if (length(valid_idx) == 0) return(tibble())
    vals <- nums[valid_idx]
    n    <- min(length(vals), length(res$col_ids))
    tibble(
      row_id  = toupper(row_letter),
      col_id  = as.integer(res$col_ids[seq_len(n)]),
      well_id = normalize_well_id(
        paste0(toupper(row_letter), res$col_ids[seq_len(n)])),
      value   = vals[seq_len(n)]
    )
  }) %>%
    mutate(
      plate_id = str_to_lower(parse_plate_id(path)),
      time_hr  = parse_time_hr(path)
    )
}

sig_label <- function(p) {
  case_when(p < 0.001 ~ "***", p < 0.01 ~ "**", p < 0.05 ~ "*",
            TRUE ~ "ns")
}
```

# LOAD DATA

## Plate export files

```{r load-plates}
proj_root <- rprojroot::find_rstudio_root_file()
data_dir  <- file.path(proj_root, "data", "volume_testing", "20280528")
fig_dir   <- file.path(proj_root, "output", "volume_testing", "20260528", "figures")
out_dir   <- file.path(proj_root, "output", "volume_testing", "20260528")

dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

plate_files <- list.files(
  data_dir,
  pattern   = "(?i)^plate-.*-T[0-9]+(?:\\.[0-9]+)?\\.txt$",
  full.names = TRUE
)

plate_raw <- map_dfr(plate_files, function(path) {
  tryCatch(
    parse_plate_export(path),
    error = function(e) {
      message("Parse error in ", basename(path), ": ", e$message)
      tibble()
    }
  )
})

str(plate_raw)
```

## Layout file

```{r load-layout}
layout_path <- file.path(data_dir, "layout.csv")

layout_raw <- read_csv(layout_path,
                       col_types    = cols(.default = "c"),
                       show_col_types = FALSE)

# Standardise column names to snake_case
names(layout_raw) <- names(layout_raw) |>
  str_to_lower() |>
  str_replace_all("[^a-z0-9]+", "_") |>
  str_replace_all("_+", "_") |>
  str_replace("_$", "")

layout_clean <- layout_raw %>%
  mutate(
    plate_id  = str_remove(str_to_lower(plate_id), "^plate-"),
    well_id   = normalize_well_id(plate_well),
    volume_ul = as.integer(family_id_group)
  )

str(layout_clean)
```

# MERGE DATA

```{r merge-data}
dat <- plate_raw %>%
  left_join(
    layout_clean %>% dplyr::select(plate_id, well_id, volume_ul),
    by = c("plate_id", "well_id")
  ) %>%
  filter(!is.na(volume_ul)) %>%
  mutate(
    volume_ul = factor(volume_ul,
                       levels = sort(unique(volume_ul)))
  )

str(dat)
```

# SUMMARY STATS

```{r summary-stats}
vol_summary <- dat %>%
  group_by(volume_ul) %>%
  summarise(
    n          = n(),
    mean_rfu   = mean(value, na.rm = TRUE),
    sd_rfu     = sd(value, na.rm = TRUE),
    se_rfu     = sd_rfu / sqrt(n),
    median_rfu = median(value, na.rm = TRUE),
    .groups    = "drop"
  )

print(vol_summary)

str(vol_summary)
```

# STATISTICAL ANALYSIS

One-way ANOVA tests for any volume effect; Tukey-adjusted pairwise comparisons
identify which volume pairs differ significantly (α = 0.05). Compact letter
display (CLD) groups volumes that are not significantly different.

## One-way ANOVA

```{r anova}
model     <- aov(value ~ volume_ul, data = dat)
anova_res <- summary(model)

print(anova_res)

str(anova_res)
```

## Tukey pairwise comparisons

```{r tukey}
tukey_res <- TukeyHSD(model)
pairs_res <- as.data.frame(tukey_res$volume_ul) %>%
  tibble::rownames_to_column("contrast") %>%
  rename(p.value = `p adj`) %>%
  mutate(sig_label = sig_label(p.value))

print(pairs_res)

str(pairs_res)
```

## Compact letter display

Groups sharing a letter are not significantly different at α = 0.05.

```{r cld}
# multcompLetters4 works directly with TukeyHSD output, so group-name
# alignment is handled internally and reliably.
cld_letters <- multcompView::multcompLetters4(model, tukey_res)$volume_ul$Letters

vols <- levels(dat$volume_ul)

cld_df <- tibble(
  volume_ul = factor(vols, levels = vols),
  .group    = trimws(cld_letters[vols])
)

print(cld_df)

str(cld_df)
```

# PLOTS

## Box Plot with Statistical Annotations

```{r boxplot-volume, fig.width = 9, fig.height = 6}
y_max   <- max(dat$value, na.rm = TRUE)
y_range <- diff(range(dat$value, na.rm = TRUE))
label_y <- y_max + y_range * 0.07

p_box <- ggplot(dat, aes(x = volume_ul, y = value, fill = volume_ul)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.55, size = 2.2) +
  geom_text(
    data        = cld_df,
    aes(x = volume_ul, y = label_y, label = .group),
    inherit.aes = FALSE,
    size        = 5,
    fontface    = "bold"
  ) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(
    x        = "Resazurin volume (µL)",
    y        = "Raw fluorescence (RFU)",
    title    = "Fluorescence by resazurin volume",
    subtitle = "Shared letter = not significantly different (Tukey, α = 0.05)"
  ) +
  theme_classic(base_size = 13) +
  theme(plot.subtitle = element_text(size = 10, colour = "grey40"))

p_box

ggsave(file.path(fig_dir, "boxplot_fluorescence_by_volume.png"),
       p_box, width = 9, height = 6)
```

## Fluorescence vs. volume (scatter with regression)

Visualises the relationship between volume and mean fluorescence to assess linearity.

```{r scatter-linearity, fig.width = 7, fig.height = 5}
vol_summary_num <- vol_summary %>%
  mutate(volume_num = as.numeric(as.character(volume_ul)))

lm_fit  <- lm(mean_rfu ~ volume_num, data = vol_summary_num)
lm_coef <- coef(lm_fit)
lm_r2   <- summary(lm_fit)$r.squared

lm_label <- sprintf(
  "y = %.3f x + %.2f\nR² = %.4f",
  lm_coef["volume_num"],
  lm_coef["(Intercept)"],
  lm_r2
)

p_scatter <- ggplot(vol_summary_num,
    aes(x = volume_num, y = mean_rfu)) +
  geom_smooth(method = "lm", colour = "steelblue", fill = "steelblue",
              alpha = 0.15, linewidth = 0.8) +
  geom_errorbar(aes(ymin = mean_rfu - se_rfu,
                    ymax = mean_rfu + se_rfu),
                width = 20, colour = "grey40") +
  geom_point(size = 3) +
  annotate("text",
           x     = min(vol_summary_num$volume_num),
           y     = max(vol_summary_num$mean_rfu),
           label = lm_label,
           hjust = 0, vjust = 1, size = 4, colour = "steelblue") +
  scale_x_continuous(breaks = vol_summary_num$volume_num) +
  labs(
    x     = "Resazurin volume (µL)",
    y     = "Mean fluorescence (RFU ± SE)",
    title = "Mean fluorescence vs. volume"
  ) +
  theme_classic(base_size = 13)

p_scatter

ggsave(file.path(fig_dir, "scatter_mean_fluorescence_by_volume.png"),
       p_scatter, width = 7, height = 5)
```

# SAVE OUTPUTS

```{r save-outputs}
write_csv(dat,         file.path(out_dir, "volume_fluorescence.csv"))
write_csv(vol_summary, file.path(out_dir, "volume_summary.csv"))
write_csv(pairs_res,   file.path(out_dir, "pairwise_stats.csv"))

message("Output files written to: ", out_dir)
```
