---
title: USDA Thermal Curves
author: "AS Huffmyer"
date: '2025'
output:
  github_document: null
  md_document: default
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 6
    toc_float: true
editor_options: 
  chunk_output_type: console
---

This script analyzes resazurin assays for the USDA thermal curve project.     

# Set up 

Set up workspace, set options, and load required packages.    
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries. 
```{r}
library(MASS) 
library(tidyverse)
library(ggplot2)
library(readxl)
library(cowplot)
library(lme4)
library(lmerTest)
library(car)
library(effects)
library(emmeans)
library(tools)
library(pracma)
```

# Resazurin metabolic rates

## Load data 

```{r}
# Set the folder path
folder_path <- "data/thermal-curve/plate-files"  

# List all txt files in the folder
file_list <- list.files(path = folder_path, pattern = "\\.txt$", full.names = TRUE, recursive=TRUE)

# Check if any files are found
if (length(file_list) == 0) {
  stop("No txt files found in the specified folder. Check the folder path or file extension.")
}

# Initialize an empty list to store processed data
data_list <- list()

# Loop through each file and load the data
for (file in file_list) {
  # Ensure file is a character string
  if (!is.character(file)) next
  
  # Extract file name without extension
  file_name <- file_path_sans_ext(basename(file))
  
  # Read the text file, skipping the first row (metadata)
  data <- read.delim(file, header = FALSE, skip = 2)
  
  # Assign column names: first column is row labels (A–H), others are 1–12
  colnames(data) <- c("Row", as.character(1:12))
  
  #remove last column
  data<-data%>%select(-ncol(data))
  
  # Convert to long format
  data_long <- data %>%
    pivot_longer(cols = -Row, names_to = "Column", values_to = "Value") %>%
    mutate(
      Column = sprintf("%02d", as.integer(Column)),  # Ensure two-digit column numbers
      Well_ID = paste0(Row, Column),                # Format well ID as "A01", "B02", etc.
      FileName = file_name,                          # Add file name
      plate = str_extract(file_name, "plate\\d+"),  # Extract "plateX"
      date = str_extract(file_name, "^\\d{8}"),     # Extract 8-digit date
      timepoint = str_extract(file_name, "T\\d+") %>% 
        str_remove("T") %>% 
        as.numeric()                                # Convert timepoint to numeric
    ) %>%
    select(FileName, Well_ID, Value, date, plate, timepoint)  # Select relevant columns
  
  # Store the processed data in the list
  data_list[[file_name]] <- data_long
}

# Print an example of processed data
head(data_list[[file_name]])


# Combine all data frames into a single data frame (optional)
combined_data <- bind_rows(data_list, .id = "Source")

# Print the first few rows of the combined data (optional)
head(combined_data)

# Rename columns
combined_data<-combined_data%>%
  rename("well"=Well_ID, resazurin_counts=`Value`)%>%
  mutate(timepoint=as.character(timepoint))

head(combined_data)
```

Load in metadata. 

```{r}
metadata<-read_xlsx(path="data/thermal-curve/metadata.xlsx")%>%
  mutate(date=as.character(date))
```

Join with data frame. 

```{r}
str(combined_data)
str(metadata)

full_data<-left_join(combined_data, metadata, by=c("date", "plate", "well"))%>%
  filter(!is.na(type))

head(full_data)
```

Load in size data. 

```{r}
size<-read_xlsx("data/thermal-curve/size.xlsx")%>%
  mutate(date=as.character(date))
```

Join with data. 

```{r}
full_data<-left_join(full_data, size, by=c("date", "plate", "well"))

full_data<-full_data%>%
  mutate(temperature=as.character(temperature.x))
```

```{r}
head(full_data)
```

## Prep the data 

Plot the raw data. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=resazurin_counts, colour=temperature, group=interaction(date, plate, well)))+
  geom_point()+
  geom_line()+
  theme_classic()
```

Calculate fluorescence at each time point normalized to the starting value at time 0. 
```{r}
full_data<-full_data%>%
  group_by(date, plate, well, temperature)%>%
  arrange(date, plate, well)%>%
  mutate(fluorescence.norm=resazurin_counts/first(resazurin_counts))
```

Plot again. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=fluorescence.norm, colour=temperature, group=interaction(date, plate, well)))+
  geom_point()+
  geom_line()+
  theme_classic()
```

View blanks

```{r}
full_data%>%
  filter(type=="blank")%>%
  ggplot(aes(x=timepoint, y=fluorescence.norm, colour=temperature, group=interaction(date, plate, well)))+
  geom_point()+
  geom_line()+
  theme_classic()
```

Calculate mean blank value at each time point. 

```{r}
blanks<-full_data%>%
  filter(type=="blank")%>%
  group_by(date, plate, temperature, timepoint)%>%
  summarise(mean_blank=mean(fluorescence.norm));blanks
```

View summarized blank data. 

```{r}
blanks%>%
  ggplot(aes(x=timepoint, y=mean_blank, colour=temperature))+
  geom_point()+
  theme_classic()
```

Subtract blank values from fluorescence values for oysters. 

```{r}
full_data<-left_join(full_data, blanks)

full_data<-full_data%>%
  filter(!type=="blank")%>%
  mutate(fluorescence.corr=fluorescence.norm-mean_blank)
```

Plot again. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=fluorescence.corr, colour=temperature, group=interaction(date, plate, well)))+
  geom_point()+
  geom_line()+
  theme_classic()
```

Size normalize data. 
```{r}
full_data<-full_data%>%
  mutate(value.mm=fluorescence.norm/length.mm)
```

Plot again. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=value.mm, colour=temperature, group=interaction(date, plate, well)))+
  geom_point()+
  geom_line()+
  theme_classic()
```

## Models    

Plot raw data. 
```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=value.mm, colour=temperature, group=interaction(date, plate, well)))+
  geom_point()+
  geom_line()+
  theme_classic()
```

```{r}
hist(full_data$value.mm)
```

Plot again. 
```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=value.mm, group=interaction(date, plate, well)))+
  facet_wrap(~temperature)+
  geom_point()+
  geom_line()+
  theme_classic()
```

Remove high value outliers. 
```{r}
full_data<-full_data%>%
  mutate(unique=paste(date, plate, temperature, well))

remove<-c("20250610 plate2 21 B02", "20250601 plate1 32 B05", "20250601 plate1 32 B02", "20250610 plate2 37 B02")

full_data<-full_data%>%
  filter(!unique %in% remove)
```

Plot again. 
```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=value.mm, group=interaction(date, plate, well)))+
  facet_wrap(~temperature)+
  geom_point()+
  geom_line()+
  theme_classic()
```

```{r}
hist(full_data$value.mm)

#remove outlier
#full_data<-full_data%>%
#  filter(value.mm<5)

hist((full_data$value.mm))
```


```{r}
model<-lmer((value.mm)^(1/3) ~ timepoint * temperature * (1|date:plate:well), data=full_data)

summary(model)
anova(model)
#rand(model)

qqPlot(residuals(model))
```

View posthoc effects. 

```{r}
emm<-emmeans(model, ~timepoint|temperature)
pairs(emm)

emm<-emmeans(model, ~temperature|timepoint)
pairs(emm)
```

## Plotting 

Plot with individual points with geom smooth lines. 

```{r}
plot1<-full_data%>%

  ggplot(aes(x=timepoint, y=value.mm, color=temperature, fill=temperature))+
  facet_grid(~date)+
  geom_point(alpha=0.5)+
  stat_smooth(aes(group=temperature), method="loess")+
  theme_classic()+
  xlab("Hour");plot1

plot1a<-full_data%>%

  ggplot(aes(x=timepoint, y=value.mm, color=temperature, fill=temperature, group=temperature))+
  geom_point(alpha=0.5)+
  stat_smooth(method="loess")+
  theme_classic()+
  xlab("Hour");plot1a

#ggsave(plot1a, filename="figures/spat-stress/metabolism_temperature.png", width=7, height=5)
```

# Total metabolic activity 

Calculate total metabolic activity as area under the curve. 

```{r}
auc_data<-full_data%>%
  mutate(time=as.numeric(timepoint))%>%
  group_by(date, plate, well, temperature) %>%
  summarise(AUC = trapz(time, value.mm))

head(auc_data)
```

Plot metabolic rates for each treatment over the temperatures. 

```{r}
plot5<-ggplot(auc_data, aes(x = temperature, y = AUC)) +
  geom_point(alpha = 0.6, position = position_jitter(height = 0.03)) +
  stat_smooth(aes(group=1)) +
  labs(
    title = "",
    y = "Metabolic Rate",
  ) +
  theme_classic();plot5

plot5a<-ggplot(auc_data, aes(x = temperature, y = AUC)) +
  geom_violin(colour="darkgray")+
  geom_point(alpha = 0.6, position = position_jitter(height = 0.01), colour="darkgray") +
  stat_smooth(aes(group=1), colour="black", method="loess") +
  geom_text(aes(x=2, y=9), label="p<0.001")+
  labs(
    title = "B",
    y = "AUC", x = "Temperature"
  ) +
  theme_classic();plot5a

ggsave(plot5a, filename="figures/thermal-curve/auc_temperatures.png", width=4, height=5)
```

Analyze effect of temperature. 

```{r}
hist(auc_data$AUC)
hist((auc_data$AUC)^(1/3))

tpc_model<-aov((AUC)^(1/3) ~ temperature, data=auc_data)

summary(tpc_model)
qqPlot(residuals(tpc_model))
```

Posthoc test.  

```{r}
emm<-emmeans(tpc_model, ~temperature)
library(multcomp)
cld(emm)
pairs(emm)
```

Calculate Q10 values. 
```{r}
# df must have columns: temp_C (°C) and rate (>0)
df<-auc_data%>%
  #filter(temperature %in% c("20", "36", "42"))%>%
  mutate(temp_C=as.numeric(temperature), rate=AUC)

# df has columns: temp_C (°C) and rate (>0)
df2 <- df %>%
  group_by(temp_C) %>%                   # average replicates at each T (optional)
  summarise(rate = mean(rate, na.rm = TRUE), .groups = "drop") %>%
  arrange(temp_C) %>%
  mutate(
    T1     = temp_C,
    T2     = lead(temp_C),
    rate_T1 = rate,
    rate_T2 = lead(rate),
    Q10    = (rate_T2 / rate_T1)^(10 / (T2 - T1))
  ) %>%
  filter(!is.na(Q10))

df2
# Yields rows for 21→26, 26→32, 32→37, 37→42, 42→45 with their Q10s
```

# Metabolic rates ~ size 

Calculate AUC prior to size normalization. 

```{r}
auc.size<-full_data%>%
  filter(temperature %in% c("21", "26"))%>%
  ungroup()%>%
  select(date, well, plate, timepoint, length.mm, fluorescence.norm)%>%
  
  mutate(time=as.numeric(timepoint))%>%
  group_by(date, well, plate, length.mm) %>%
  dplyr::summarise(AUC = trapz(time, fluorescence.norm));auc.size
  
```

```{r}
size_plot<-auc.size%>%
  
  ggplot(data=., aes(x=length.mm, y=AUC, group=paste(date, well, plate)))+
  geom_point(colour="darkgray")+
  geom_smooth(aes(group=1), method="lm", colour="black")+
  xlab("Length (mm)")+
  ylab("AUC")+
  theme_classic();size_plot
```

Export data for aggregate size analysis. 

```{r}
auc.size%>%
  write.csv("output/size/thermal_curve_size.csv")
```



