---
title: PolyIC resazurin testing for batch 2 seed 
author: "AS Huffmyer"
date: '2024'
output:
  github_document: null
  md_document: default
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 6
    toc_float: true
editor_options: 
  chunk_output_type: console
---

This script analyzes resazurin assays for the PolyIC seed project. Analyzed with the FLx800 plate reader instrument.    

# Set up 

Set up workspace, set options, and load required packages.    
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries. 
```{r}
library(MASS) 
library(tidyverse)
library(ggplot2)
library(readxl)
library(cowplot)
library(lme4)
library(lmerTest)
library(car)
library(effects)
library(emmeans)
library(tools)
library(pracma)
```

# Resazurin metabolic rates

## Load data 

```{r}
# Set the folder path
folder_path <- "data/spat-stress/plate-files"  

# List all txt files in the folder
file_list <- list.files(path = folder_path, pattern = "\\.txt$", full.names = TRUE)

# Check if any files are found
if (length(file_list) == 0) {
  stop("No txt files found in the specified folder. Check the folder path or file extension.")
}

# Initialize an empty list to store processed data
data_list <- list()

# Loop through each file and load the data
for (file in file_list) {
  # Ensure file is a character string
  if (!is.character(file)) next
  
  # Extract file name without extension
  file_name <- file_path_sans_ext(basename(file))
  
  # Read the text file, skipping the first row (metadata)
  data <- read.delim(file, header = FALSE, skip = 2)
  
  # Assign column names: first column is row labels (A–H), others are 1–12
  colnames(data) <- c("Row", as.character(1:13))
  
  #remove last column
  data<-data%>%select(-ncol(data))
  
  # Convert to long format
  data_long <- data %>%
    pivot_longer(cols = -Row, names_to = "Column", values_to = "Value") %>%
    mutate(
      Column = sprintf("%02d", as.integer(Column)),  # Ensure two-digit column numbers
      Well_ID = paste0(Row, Column),                # Format well ID as "A01", "B02", etc.
      FileName = file_name,                          # Add file name
      plate = str_extract(file_name, "plate\\d+"),  # Extract "plateX"
      date = str_extract(file_name, "^\\d{8}"),     # Extract 8-digit date
      timepoint = str_extract(file_name, "T\\d+") %>% 
        str_remove("T") %>% 
        as.numeric()                                # Convert timepoint to numeric
    ) %>%
    select(FileName, Well_ID, Value, plate, date, timepoint)  # Select relevant columns
  
  # Store the processed data in the list
  data_list[[file_name]] <- data_long
}

# Print an example of processed data
head(data_list[[file_name]])


# Combine all data frames into a single data frame (optional)
combined_data <- bind_rows(data_list, .id = "Source")

# Print the first few rows of the combined data (optional)
head(combined_data)

# Rename columns
combined_data<-combined_data%>%
  rename("well"=Well_ID, resazurin_counts=`Value`)%>%
  mutate(timepoint=as.character(timepoint))

head(combined_data)
```

Load in metadata. 

```{r}
metadata<-read_xlsx(path="data/spat-stress/metadata_spat.xlsx")%>%
  mutate(date=as.character(date))
```

Join with data frame and remove any wells that did not have samples. 

```{r}
str(combined_data)
str(metadata)

full_data<-left_join(combined_data, metadata, by=c("date", "plate", "well"))%>%
  filter(!is.na(type))

head(full_data)
```

Load in size data. 

```{r}
size<-read_csv("data/spat-stress/size-spat.csv")%>%
  mutate(date=as.character(date))%>%select(!notes)
```

Join with data. 

```{r}
full_data<-left_join(full_data, size, by=c("date", "plate", "well"))
```

## Prep the data 

Plot the raw data. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=resazurin_counts, colour=temperature, group=interaction(date, plate, well)))+
  facet_wrap(~date)+
  geom_point()+
  geom_line()+
  scale_colour_manual(values=c("20C"="cyan4", "36C"="purple2", "38C"="pink", "40C"="orange", "42C"="darkred"))+
  theme_classic()
```

Calculate blank corrected fluorescence at each time point normalized to the starting value at time 0. 
```{r}
full_data<-full_data%>%
  group_by(date, plate, well, temperature, spat_treatment, tank, holding_tank)%>%
  arrange(date, plate, well)%>%
  mutate(fluorescence.norm=resazurin_counts/first(resazurin_counts))
```

Plot again. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=fluorescence.norm, colour=temperature, group=interaction(date, plate, well)))+
  facet_wrap(~date)+
  geom_point()+
  geom_line()+
  scale_colour_manual(values=c("20C"="cyan4", "36C"="purple2", "38C"="pink", "40C"="orange", "42C"="darkred"))+
  theme_classic()
```

View blanks

```{r}
full_data%>%
  filter(type=="blank")%>%
  ggplot(aes(x=timepoint, y=fluorescence.norm, colour=temperature, group=interaction(date, plate, well)))+
  facet_wrap(~date)+
  geom_point()+
  scale_colour_manual(values=c("20C"="cyan4", "36C"="purple2", "38C"="pink", "40C"="orange", "42C"="darkred"))+
  geom_line()+
  theme_classic()
```

Calculate mean blank value at each time point. 

```{r}
blanks<-full_data%>%
  filter(type=="blank")%>%
  group_by(date, plate, temperature, timepoint)%>%
  summarise(mean_blank=mean(fluorescence.norm));blanks
```

View summarized blank data. 

```{r}
blanks%>%
  ggplot(aes(x=timepoint, y=mean_blank, colour=temperature))+
  facet_wrap(~date)+
  geom_point()+
  scale_colour_manual(values=c("20C"="cyan4", "36C"="purple2", "38C"="pink", "40C"="orange", "42C"="darkred"))+
  theme_classic()
```

Subtract blank values from fluorescence values for oysters. 

```{r}
full_data<-left_join(full_data, blanks)

full_data<-full_data%>%
  filter(!type=="blank")%>%
  mutate(fluorescence.corr=resazurin_counts-mean_blank)
```

Plot again. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=fluorescence.corr, colour=temperature, group=interaction(date, plate, well)))+
  facet_wrap(~date)+
  geom_point()+
  geom_line()+
  scale_colour_manual(values=c("20C"="cyan4", "36C"="purple2", "38C"="pink", "40C"="orange", "42C"="darkred"))+
  theme_classic()
```

Size normalize data. 
```{r}
full_data<-full_data%>%
  mutate(value.mm=fluorescence.norm/length.mm)
```

Plot again. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=value.mm, colour=temperature, group=interaction(date, plate, well)))+
  facet_wrap(~date)+
  geom_point()+
  geom_line()+
  scale_colour_manual(values=c("20C"="cyan4", "36C"="purple2", "38C"="pink", "40C"="orange", "42C"="darkred"))+
  theme_classic()
```

Remove unnecessary columns. 

```{r}
full_data<-full_data%>%
  select(!resazurin_counts)%>%
  select(!FileName)%>%
  select(!type)%>%
  select(!fluorescence.corr)%>%
  select(!mean_blank)
```

## Models    

Plot raw data. 
```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=value.mm, colour=temperature, group=interaction(date, plate, well)))+
  facet_wrap(~date*spat_treatment)+
  geom_point()+
  scale_colour_manual(values=c("20C"="cyan4", "36C"="purple2", "38C"="pink", "40C"="orange", "42C"="darkred"))+
  geom_line()+
  theme_classic()
```

```{r}
hist(full_data$value.mm)
hist(full_data$length.mm)
```

Plot again. 
```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=value.mm, group=interaction(date, plate, well)))+
  facet_wrap(~temperature)+
  geom_point()+
  geom_line()+
  theme_classic()
```

```{r}
model<-lmer((value.mm)^(1/3) ~ timepoint * temperature * (1|date:plate:well) + (1|holding_tank) + (1|spat_treatment), data=full_data)

summary(model)
anova(model)
#rand(model)

qqPlot(residuals(model))
```

View posthoc effects. 

```{r}
emm<-emmeans(model, ~timepoint|temperature)
pairs(emm)

emm<-emmeans(model, ~temperature|timepoint)
pairs(emm)
```

## Plotting 

Plot with individual points with geom smooth lines. 

```{r}
plot1<-full_data%>%

  ggplot(aes(x=timepoint, y=value.mm, color=temperature, fill=temperature))+
  facet_grid(~date)+
  geom_point(alpha=0.5)+
  stat_smooth(aes(group=temperature))+
  scale_colour_manual(values=c("20C"="cyan4", "36C"="purple2", "38C"="pink", "40C"="orange", "42C"="darkred"))+
  scale_fill_manual(values=c("20C"="cyan4", "36C"="purple2", "38C"="pink", "40C"="orange", "42C"="darkred"))+
  theme_classic()+
  xlab("Hour");plot1

plot1a<-full_data%>%

  ggplot(aes(x=timepoint, y=value.mm, color=temperature, fill=temperature, group=temperature))+
  geom_point(alpha=0.5)+
  stat_smooth(method="loess")+
    scale_colour_manual(values=c("20C"="cyan4", "36C"="purple2", "38C"="pink", "40C"="orange", "42C"="darkred"))+
  scale_fill_manual(values=c("20C"="cyan4", "36C"="purple2", "38C"="pink", "40C"="orange", "42C"="darkred"))+
  theme_classic()+
  xlab("Hour");plot1a

ggsave(plot1a, filename="figures/spat-stress/metabolism_temperature.png", width=7, height=5)
```

# Survival 

## Load data 

```{r}
surv<-read_csv("data/spat-stress/survival-spat.csv")%>%mutate(date=as.character(date))%>%select(!notes)
```

Merge with existing data. 

```{r}
full_data<-left_join(full_data, surv)
```

## Analyze survival 

Analyze with binomial logistic regression for effects of temperature and spat treatment with tank as random effect. 

0 = alive
1 = mortality 

```{r}
surv_model_data<-full_data%>%
  filter(timepoint=="4")
  
surv_model<-glmer(mortality ~ temperature * spat_treatment + (1|holding_tank) + (1|spat_treatment:tank), family = binomial(link = "logit"), data=surv_model_data)

summary(surv_model)
Anova(surv_model)
```

Plot model effects. 

```{r}
plot(Effect(c("spat_treatment", "temperature"), surv_model, residuals=TRUE))
plot(Effect(c("temperature"), surv_model, residuals=TRUE))
plot(Effect(c("spat_treatment"), surv_model, residuals=TRUE))
```

Generate predictive mortality. 

```{r}
# Generate predicted probabilities
surv_model_data$predicted_mortality <- predict(surv_model, type = "response")

# Plot
plot2<-ggplot(surv_model_data, aes(x = temperature, y = predicted_mortality, color = spat_treatment, fill = spat_treatment, group=spat_treatment)) +
  geom_point(aes(y = mortality), alpha = 0.6, position = position_jitter(height = 0.03)) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), se = TRUE) +
  scale_colour_manual(values=c("darkgray", "darkred"))+
  scale_fill_manual(values=c("darkgray", "darkred"))+
  labs(
    title = "",
    y = "Probability of Mortality",
  ) +
  theme_classic();plot2

ggsave(plot2, filename="figures/spat-stress/mortality_treatment.png", width=6, height=5)

# Plot
plot2b<-ggplot(surv_model_data, aes(x = temperature, y = predicted_mortality)) +
  geom_point(aes(y = mortality), alpha = 0.6, position = position_jitter(height = 0.03), colour="darkgray") +
  geom_smooth(aes(group=1), method = "glm", method.args = list(family = "binomial"), se = TRUE, colour="black", fill="darkgray") +
  #scale_colour_manual(values=c("darkgray", "darkred"))+
  #scale_fill_manual(values=c("darkgray", "darkred"))+
  labs(
    title = "",
    y = "Probability of Mortality",
  ) +
  theme_classic();plot2b


ggsave(plot2b, filename="figures/spat-stress/mortality_all.png", width=6, height=5)

# Plot
plot2a<-ggplot(surv_model_data, aes(x = spat_treatment, y = predicted_mortality, color = temperature, fill = temperature, group=temperature)) +
  geom_point(aes(y = mortality), alpha = 0.6, position = position_jitter(height = 0.03)) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE) +
    scale_colour_manual(values=c("20C"="cyan4", "36C"="purple2", "38C"="pink", "40C"="orange", "42C"="darkred"))+
  scale_fill_manual(values=c("20C"="cyan4", "36C"="purple2", "38C"="pink", "40C"="orange", "42C"="darkred"))+
  labs(
    title = "",
    y = "Probability of Mortality",
  ) +
  theme_classic();plot2a

ggsave(plot2a, filename="figures/spat-stress/mortality_temperature.png", width=6, height=5)
```

Plot individual tanks. 

```{r}
# Plot
plot3<-ggplot(surv_model_data, aes(x = temperature, y = predicted_mortality, color = tank, fill = tank, group=tank)) +
  geom_point(aes(y = mortality), alpha = 0.6, position = position_jitter(height = 0.03)) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), se = TRUE) +
  labs(
    title = "",
    y = "Probability of Mortality",
  ) +
  theme_classic();plot3
```


# Metabolic rates ~ mortality 

Plot metabolic rates at 42°C (treatment with highest mortality) for those that lived and those that died in each treatment. 

```{r}
data_42C<-full_data%>%
  filter(temperature=="42C")%>%
  mutate(temperature_num = as.numeric(str_extract(temperature, "\\d+")))

str(data_42C)
```

Analyze the effect of mortality on metabolic rates. 
```{r}
hist(data_42C$value.mm)
data_42C<-data_42C%>%
  filter(value.mm<4) #remove outliers

mortality_model<-lmer((value.mm)^(1/3) ~ timepoint * mortality + (1|spat_treatment) + (1|holding_tank) + (1|spat_treatment:tank), data=data_42C)

summary(mortality_model)
anova(mortality_model)

qqPlot(residuals(mortality_model))
```

Time and mortality is significant 

Plot by mortality. 

```{r}
plot4a<-ggplot(data_42C, aes(x = timepoint, y = value.mm, color = as.factor(mortality), fill = as.factor(mortality))) +
  geom_point(alpha = 0.6, position = position_jitter(height = 0.01)) +
  stat_smooth(aes(group=mortality), method="lm") +
  scale_colour_manual(values=c("black", "darkgray"), labels=c("alive", "dead"), name="Mortality")+
  scale_fill_manual(values=c("black", "darkgray"), labels=c("alive", "dead"), name="Mortality")+
  geom_text(aes(x=3, y=4), label="p<0.001", colour="black")+
  labs(
    title = "C",
    y = expression(paste("Fold Change Fluorescence mm" ^-1)),
    x = "Hours"
  ) +
  theme_classic();plot4a

ggsave(plot4a, filename="figures/spat-stress/mortality_metabolic_rates_42C.png", width=5, height=5)
```

Calculate AUC for alive and dead oysters and plot. 

```{r}
auc_data<-data_42C%>%
  mutate(time=as.numeric(timepoint))%>%
  group_by(date, plate, well, temperature, mortality) %>%
  summarise(AUC = trapz(time, value.mm))%>%
  mutate(mortality=if_else(mortality=="0", "alive", "dead"))

head(auc_data)
```

```{r}
plot4b<-auc_data%>%
  group_by(mortality)%>%
  summarise(mean=mean(AUC), se=sd(AUC)/sqrt(length(AUC)))%>%
  
  ggplot(., aes(x = mortality, y = mean, color = mortality, fill = mortality)) +
  geom_point() +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=0.1) +
  scale_colour_manual(values=c("black", "darkgray"), name="Mortality")+
 scale_fill_manual(values=c("black", "darkgray"), name="Mortality")+
  geom_text(aes(x=1.5, y=4.5), label="p<0.001", colour="black")+
  ylim(0,5)+
  labs(
    title = "D",
    y = expression(paste("AUC")),
    x = "Mortality"
  ) +
  theme_classic()+
  theme(legend.position="none");plot4b

ggsave(plot4b, filename="figures/spat-stress/mortality_metabolic_rates_42C_mean.png", width=3, height=4)
```

```{r}
hist(auc_data$AUC)

tpc_model<-aov((AUC)^(1/3) ~ mortality, data=auc_data)

summary(tpc_model)
anova(tpc_model)
qqPlot(residuals(tpc_model))
```

# Metabolic rates across thermal range 

Calculate total metabolic activity as area under the curve.  

```{r}
auc_data<-full_data%>%
  mutate(time=as.numeric(timepoint))%>%
  group_by(date, plate, well, temperature, spat_treatment, tank, holding_tank) %>%
  summarise(AUC = trapz(time, value.mm))%>%
  mutate(temperature = str_replace_all(temperature, "C", ""))

head(auc_data)
```

Plot metabolic rates over the temperatures. 

```{r}
plot5a<-ggplot(auc_data, aes(x = temperature, y = AUC)) +
  geom_violin(colour="darkgray")+
  geom_point(alpha = 0.6, position = position_jitter(height = 0.01), colour="darkgray") +
  stat_smooth(aes(group=1), colour="black", method="loess") +
  geom_text(aes(x=3, y=15), label="p<0.001")+
  labs(
    title = "A",
    y = "AUC", x = "Temperature"
  ) +
  theme_classic();plot5a

ggsave(plot5a, filename="figures/spat-stress/auc_spat.png", width=4, height=5)
```

Analyze effect of temperature. 

```{r}
hist(auc_data$AUC)

tpc_model<-lmer((AUC)^(1/3) ~ temperature + (1|holding_tank) + (1|spat_treatment) + (1|spat_treatment:tank), data=auc_data)

summary(tpc_model)
anova(tpc_model)
qqPlot(residuals(tpc_model))
```

Posthoc test.  

```{r}
emm<-emmeans(tpc_model, ~temperature)
pairs(emm)
```

Calculate Q10 values. 
```{r}
# df must have columns: temp_C (°C) and rate (>0)
df<-auc_data%>%
  filter(temperature %in% c("20", "36", "42"))%>%
  mutate(temp_C=as.numeric(temperature), rate=AUC)

# df has columns: temp_C (°C) and rate (>0)
df2 <- df %>%
  group_by(temp_C) %>%                   # average replicates at each T (optional)
  summarise(rate = mean(rate, na.rm = TRUE), .groups = "drop") %>%
  arrange(temp_C) %>%
  mutate(
    T1     = temp_C,
    T2     = lead(temp_C),
    rate_T1 = rate,
    rate_T2 = lead(rate),
    Q10    = (rate_T2 / rate_T1)^(10 / (T2 - T1))
  ) %>%
  filter(!is.na(Q10))

df2
# Yields rows for 20→36, 36→42 with their Q10s
```

# Metabolic rates ~ size 

Plot fluorescence data by size.  

```{r}
plot6a<-full_data%>%
  filter(timepoint=="4")%>%
  filter(temperature %in% c("36C", "38C"))%>%
  
  ggplot(., aes(x = length.mm, y = fluorescence.norm)) +
  geom_point(alpha = 0.6, position = position_jitter(height = 0.01), colour="darkgray") +
  stat_smooth(aes(group=1), colour="black", method="lm") +
  geom_text(aes(x=4.5, y=35), label="p=0.002")+
  labs(
    title = "A",
    y = "Fold Change in Fluorescence", x = "Length (mm)"
  ) +
  theme_classic();plot6a

ggsave(plot6a, filename="figures/spat-stress/metabolism_size.png", width=5, height=5)
```

```{r}
size_model<-full_data%>%
  filter(timepoint=="4")%>%
  filter(temperature %in% c("36C", "38C"))%>%
  
  lmer((fluorescence.norm)^(1/3) ~ scale(length.mm) + (1|temperature), data=.)

summary(size_model)
anova(size_model)
qqPlot(residuals(size_model))
```



