---
title: Oxygen rate extraction 
author: "AS Huffmyer"
date: '2025'
output:
  github_document: null
  md_document: default
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 6
    toc_float: true
editor_options: 
  chunk_output_type: console
---

This script extracts respiration rates from oxygen data collected to correlate with resazurin response.      

# Set up 

Set up workspace, set options, and load required packages.    
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r, warning=FALSE, message=FALSE}
## install packages if you dont already have them in your library
if ("devtools" %in% rownames(installed.packages()) == 'FALSE') install.packages('devtools') 
library(devtools)
if ("segmented" %in% rownames(installed.packages()) == 'FALSE') install.packages('segmented') 
if ("plotrix" %in% rownames(installed.packages()) == 'FALSE') install.packages('plotrix') 
if ("gridExtra" %in% rownames(installed.packages()) == 'FALSE') install.packages('gridExtra') 
if ("LoLinR" %in% rownames(installed.packages()) == 'FALSE') install_github('colin-olito/LoLinR') 
if ("lubridate" %in% rownames(installed.packages()) == 'FALSE') install.packages('lubridate') 
if ("chron" %in% rownames(installed.packages()) == 'FALSE') install.packages('chron') 
if ("plyr" %in% rownames(installed.packages()) == 'FALSE') install.packages('plyr') 
if ("dplyr" %in% rownames(installed.packages()) == 'FALSE') install.packages('dplyr') 
if ("stringr" %in% rownames(installed.packages()) == 'FALSE') install.packages('stringr') 
if ("Rmisc" %in% rownames(installed.packages()) == 'FALSE') install.packages('Rmisc') 
if ("respR" %in% rownames(installed.packages()) == 'FALSE') install.packages('respR') 


#load packages
library("ggplot2")
library("segmented")
library("plotrix")
library("gridExtra")
library("LoLinR")
library("lubridate")
library("chron")
library('plyr')
library('dplyr')
library('stringr')
library('Rmisc')
library('respR')
library('readxl')
library('tidyverse')
```

## Read in files  

Set the path of all respirometry files within the R project.   

```{r, warning=FALSE, message=FALSE}
path.p<-"data/oxygen-correlation/oxygen-files" #location of files
```

Bring in the file names.
```{r, warning=FALSE, message=FALSE}
# bring in the respiration file names
file.names<-basename(list.files(path = path.p, pattern = "csv$", recursive = TRUE)) 
```

## Extract respiration rates  

Load in the sample information file. 

```{r, warning=FALSE, message=FALSE}
Sample.Info<-read_excel("data/oxygen-correlation/oxygen_samples.xlsx") 
```

Run loop to extract slopes from respiration data. 

Oxygen is in mmol/L concentration in original files. 

```{r}
# Set paths
input_dir <- "data/oxygen-correlation/oxygen-files"
output_dir <- "output/oxygen-correlation/RespirationPlots"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# Get list of all Excel files
files <- list.files(input_dir, pattern = "\\.xlsx$", full.names = TRUE);files

# Initialize summary table
summary_df <- data.frame(Plate = character(),
                         Well = character(),
                         Slope = numeric(),
                         Intercept = numeric(),
                         stringsAsFactors = FALSE)

# Loop through each file
for (file in files) {
  # Get plate name from file name (e.g., "plate1.xlsx" -> "plate1")
  plate_name <- tools::file_path_sans_ext(basename(file))
  
  # Read data
  data <- read_xlsx(file, skip = 50)
  data$Time.Min <- seq.int(from = 0, to = ((nrow(data) * 0.25) - 0.25), by = 0.25)
  
  # Subset wells (columns 53–77) and rename
  data.sub <- data[, 53:77]
  names(data.sub) <- gsub("\\s*\\[Oxygen\\]", "", names(data.sub))
  
  # Add Time column to subset
  data.sub$Time.Min <- data$Time.Min
  
  # Loop over each well (column)
  for (well_name in names(data.sub)[names(data.sub) != "Time.Min"]) {
    x <- data.sub$Time.Min
    y <- as.numeric(data.sub[[well_name]])
    
    # Skip if y is all NA
    if (all(is.na(y))) next
    
    # Run localized regression
    model <- rankLocReg(xall = x, yall = y,
                        alpha = 0.4, method = "pc", verbose = FALSE)
    
    # Extract slope and intercept from model
    best_idx <- which.min(model$allRegs$ciRange) #This picks the regression with the narrowest confidence interval, which is what the "pc" (percentile confidence interval) method is designed to optimize.
    slope <- model$allRegs$b1[best_idx]
    intercept <- model$allRegs$b0[best_idx]
    
    # Save plot
    pdf(file.path(output_dir, paste0(plate_name, "_", well_name, "_regression_trunc.pdf")))
    plot(model)
    dev.off()
    
    # Append to summary table
    summary_df <- rbind(summary_df,
                        data.frame(Plate = plate_name,
                                   Well = well_name,
                                   Slope = slope,
                                   Intercept = intercept,
                                   stringsAsFactors = FALSE))
  }
}

# View or save summary
head(summary_df)
```

Merge in data from the metadata file. 

```{r}
metadata<-read_xlsx("data/oxygen-correlation/oxygen_samples.xlsx")

summary_df<-summary_df%>%
  rename(plate=Plate, well=Well, slope=Slope, intercept=Intercept)

oxygen<-left_join(summary_df, metadata)

head(oxygen)
```

Conduct calculations. 

```{r}
oxygen$R.mmol.L.min<-oxygen$slope
oxygen$volume<-0.0017 #volume in L
oxygen$R.mmol.min <- oxygen$R.mmol.L.min * oxygen$volume #calculate total oxygen
```

View values for clean vs dirty vs shell only blanks. 
```{r}
plot1<-oxygen%>%
  
  ggplot(aes(x=notes, y=R.mmol.min))+
  geom_boxplot()+
  geom_point()+
  theme_classic(); plot1
```

Similar between clean and dirty water blanks. Shell and "samples" are similar - confirming that with the exception of potentially one sample, most oysters were likely empty shells today.  

Subtract clean water blanks from biological samples.  

```{r}
blank_data <- subset(oxygen, c(type == "blank" & notes=="clean")) #subset to blank data only

mean(blank_data$R.mmol.min, na.rm=TRUE) #view overall mean

resp.blnk <- aggregate(R.mmol.min ~ plate, data=blank_data, mean) #calculate average blank during light for each run

colnames(resp.blnk)[colnames(resp.blnk) == 'R.mmol.min'] <- 'R.Blank.mmol.min' #rename to specify blank for R

oxygen <- full_join(oxygen, resp.blnk) #add R blanks to dataframe

#subtract blanks to get a corrected value for each sample 
oxygen$R.mmol.min.corr<-oxygen$R.mmol.min-oxygen$R.Blank.mmol.min #subtract R blanks

plot(oxygen$R.mmol.min.corr)
dev.off()
```

Normalize to oyster length.  

```{r, warning=FALSE, message=FALSE}
#read in length 
size<-read_xlsx("data/oxygen-correlation/size-images/size.xlsx")

oxygen.bio <- oxygen %>% filter(type == "sample") #isolate only biological samples and drop unused factor levels 
oxygen.bio <- droplevels(oxygen.bio) #drop unused factor levels

#merge in size data 
oxygen.bio<-left_join(oxygen.bio, size)

#respiration
oxygen.bio$R.mmol.mm.min <- oxygen.bio$R.mmol.min.corr/oxygen.bio$length.mm #calculate oxygen per organism 
oxygen.bio$R.umol.mm.min <- oxygen.bio$R.mmol.mm.min*1000 #calculate micromoles 
```

Plot values.  
```{r}
plot(oxygen.bio$R.umol.mm.min)
```

Plot length against total oxygen. 

```{r}
plot(oxygen.bio$length.mm, oxygen.bio$R.mmol.min)
```

Save extracted values. 

```{r}
oxygen.bio%>%
  select(plate, well, oyster, length.mm, R.mmol.min, R.mmol.mm.min, R.umol.mm.min)%>%
  write_csv(., file="output/oxygen-correlation/extracted_oxygen_rates.csv")
```



