---
title: PolyIC resazurin testing for batch 2 seed 
author: "AS Huffmyer"
date: '2024'
output:
  github_document: null
  md_document: default
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 6
    toc_float: true
editor_options: 
  chunk_output_type: console
---

This script analyzes resazurin assays tracking individual metabolism in ambient and high temperatures in the USDA families project.   

# Set up 

Set up workspace, set options, and load required packages.    
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries. 
```{r}
library(MASS) 
library(tidyverse)
library(ggplot2)
library(readxl)
library(cowplot)
library(lme4)
library(lmerTest)
library(car)
library(effects)
library(emmeans)
library(tools)
```

# Fluorescence 

## Load data 

```{r}
# Set the folder path
folder_path <- "data/usda-families/fluorescence"  

# List all txt files in the folder
file_list <- list.files(path = folder_path, pattern = "\\.txt$", full.names = TRUE, recursive=TRUE)

# Check if any files are found
if (length(file_list) == 0) {
  stop("No txt files found in the specified folder. Check the folder path or file extension.")
}

# Initialize an empty list to store processed data
data_list <- list()

# Loop through each file and load the data
for (file in file_list) {
  # Ensure file is a character string
  if (!is.character(file)) next
  
  # Extract file name without extension
  file_name <- file_path_sans_ext(basename(file))
  
  # Read the text file, skipping the first row (metadata)
  data <- read.delim(file, header = FALSE, skip = 40)
  
  # Assign column names: first column is row labels (A–H), others are 1–12
  colnames(data) <- c("Row", as.character(1:13))
  
  #remove last column
  data<-data%>%select(-ncol(data))
  
  # Convert to long format
  data_long <- data %>%
    pivot_longer(cols = -Row, names_to = "Column", values_to = "Value") %>%
    mutate(
      Column = sprintf("%02d", as.integer(Column)),  # Ensure two-digit column numbers
      Well_ID = paste0(Row, Column),                # Format well ID as "A01", "B02", etc.
      FileName = file_name,                          # Add file name
      date = str_extract(file_name, "^\\d{8}"),     # Extract 8-digit date
      timepoint = str_extract(file_name, "T\\d+") %>% 
        str_remove("T") %>% 
        as.numeric()                                # Convert timepoint to numeric
    ) %>%
    select(FileName, Well_ID, Value, date, timepoint)  # Select relevant columns
  
  # Store the processed data in the list
  data_list[[file_name]] <- data_long
}

# Print an example of processed data
head(data_list[[file_name]])


# Combine all data frames into a single data frame (optional)
combined_data <- bind_rows(data_list, .id = "Source")

# Print the first few rows of the combined data (optional)
head(combined_data)

# Rename columns
combined_data<-combined_data%>%
  rename("well"=Well_ID, resazurin_counts=`Value`)%>%
  mutate(timepoint=as.character(timepoint))

head(combined_data)
```

Load in metadata. 

```{r}
metadata<-read_xlsx(path="data/usda-families/metadata/sample_metadata.xlsx")%>%
  mutate(date=as.character(date))

metadata_temp<-read_xlsx(path="data/usda-families/metadata/temperature_metadata.xlsx")%>%
  mutate(date=as.character(date))%>%mutate(timepoint=as.character(timepoint))
```

Join with data frame and remove any wells that did not have samples. 

```{r}
str(combined_data)
str(metadata)
str(metadata_temp)

full_data<-left_join(combined_data, metadata, by=c("date", "well"))%>%
  filter(!is.na(type))

full_data<-left_join(full_data, metadata_temp, by=c("date", "timepoint"))%>%
  filter(!is.na(type))

head(full_data)
```

Load in size data. 

```{r}
size<-read_xlsx("data/usda-families/size/size.xlsx")%>%
  mutate(date=as.character(date))
```

Join with data. 

```{r}
str(full_data)
str(size)

full_data<-left_join(full_data, size, by=c("date", "sample"))
```

## Prep the data 

Remove sample that was an empty shell. 

```{r}
full_data<-full_data%>%
  filter(!c(date=="20250527" & sample=="B5"))%>%
  filter(!c(date=="20250603" & sample=="B1"))%>%
  filter(!c(date=="20250603" & sample=="C3"))
```

Plot the raw data. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=resazurin_counts, colour=temperature, group=interaction(date, well)))+
  facet_wrap(~family*date)+
  geom_point()+
  geom_line()+
  scale_colour_manual(values=c("darkblue", "darkred"))+
  theme_classic()
```

Calculate size normalized fluorescence at each time point normalized to the starting value at time 0. 
```{r}
full_data<-full_data%>%
  group_by(date, well, family)%>%
  arrange(date, well)%>%
  mutate(fluorescence.norm=resazurin_counts/first(resazurin_counts))
```

Plot again. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=fluorescence.norm, colour=temperature, group=interaction(date, well)))+
  facet_wrap(~family*date)+
  geom_point()+
  geom_line()+
  scale_colour_manual(values=c("darkblue", "darkred"))+
  theme_classic()
```

View blanks

```{r}
full_data%>%
  filter(type=="blank")%>%
  ggplot(aes(x=timepoint, y=fluorescence.norm, colour=temperature, group=interaction(date, well)))+
  facet_wrap(~date)+
  geom_point()+
  scale_colour_manual(values=c("darkblue", "darkred"))+
  geom_line()+
  theme_classic()
```

Calculate mean change in blank at each time point. 

```{r}
blanks<-full_data%>%
  filter(type=="blank")%>%
  group_by(date, timepoint)%>%
  summarise(mean_blank=mean(fluorescence.norm));blanks
```

View summarized blank data. 

```{r}
blanks%>%
  ggplot(aes(x=timepoint, y=mean_blank))+
  facet_wrap(~date)+
  geom_point()+
  theme_classic()
```

Subtract blank values from fluorescence values for oysters. 

```{r}
full_data<-left_join(full_data, blanks)

full_data<-full_data%>%
  filter(!type=="blank")%>%
  mutate(fluorescence.corr=fluorescence.norm-mean_blank)
```

Plot again. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=fluorescence.corr, colour=temperature, group=interaction(date, well)))+
  facet_wrap(~family*date)+
  geom_point()+
  geom_line()+
  scale_colour_manual(values=c("darkblue", "darkred"))+
  theme_classic()
```

Size normalize data. I am going to use area first. 
```{r}
#normalize by area
full_data<-full_data%>%
  mutate(fluorescence.corr.mm2=fluorescence.corr/area)

#normalize by length=
full_data<-full_data%>%
  mutate(fluorescence.corr.mm=fluorescence.corr/length.mm)
```

Plot again. 

Plot by area normalized. 
```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=fluorescence.corr.mm2, colour=temperature, group=interaction(date, well)))+
  facet_wrap(~family*date)+
  geom_point()+
  geom_line()+
  scale_colour_manual(values=c("darkblue", "darkred"))+
  theme_classic()
```

Plot by length
```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=fluorescence.corr.mm, colour=temperature, group=interaction(date, well)))+
  facet_wrap(~family*date)+
  geom_point()+
  geom_line()+
  scale_colour_manual(values=c("darkblue", "darkred"))+
  theme_classic()
```

We will use area moving forward. Both seem to give us very similar answers and area will be more representative of total size.  

Plot again. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=fluorescence.corr.mm2, colour=temperature, group=interaction(date, well)))+
  facet_wrap(~family*date)+
  geom_point()+
  geom_line()+
  scale_colour_manual(values=c("darkblue", "darkred"))+
  theme_classic()
```

Remove unnecessary columns. 

```{r}
str(full_data)

full_data<-full_data%>%
  select(!Source)%>%
  select(!FileName)%>%
  select(!resazurin_counts)%>%
  select(!type)%>%
  select(!mean_blank)%>%
  select(!area)%>%
  select(!length.mm)%>%
  select(!fluorescence.corr)%>%
  select(!fluorescence.norm)

str(full_data)

full_data<-full_data%>%
  rename(value=fluorescence.corr.mm2)
```


## Models and Plots    

Plot raw data. 
```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=value, colour=family, group=interaction(date, family, well)))+
  facet_wrap(~date)+
  geom_point()+
  geom_vline(xintercept=4, linetype="dashed")+
  scale_colour_manual(values=c("A"="cyan4", "B"="purple2", "C"="pink", "D"="orange", "E"="darkred"))+
  geom_line()+
  theme_classic()
```

```{r}
plot1<-full_data%>%
  ggplot(aes(x=timepoint, y=value, colour=temperature, group=interaction(date, well)))+
  facet_wrap(~family)+
  geom_point()+
  geom_vline(xintercept=4, linetype="dashed")+
  scale_colour_manual(values=c("darkblue", "darkred"))+
  geom_line()+
  theme_classic();plot1

ggsave(plot1, filename="figures/usda-families/fl_family_trajectories.png", width=8, height=8, units="in")
```

## Calculate change due to temperature exposure   

Add a column for time of exposure. 

```{r}
full_data<-full_data%>%
  mutate(time_exposure=case_when(
    timepoint=="0" ~ 0,
    timepoint=="1" ~ 1,
    timepoint=="2" ~ 2,
    timepoint=="3" ~ 3,
    timepoint=="4" ~ 1,
    timepoint=="5" ~ 2,
    timepoint=="6" ~ 3
  ))

#replicate data from time point 3 as time of exposure 0 for 42C treatment and time of exposure 3 for 18C treatment

str(full_data)

# Create a copy of the hour 3 rows
hour3_rows <- full_data %>% 
  filter(timepoint == "3")

# Create ambient treatment version
ambient_rows <- hour3_rows %>%
  mutate(temperature = "18C", timepoint = "3", time_exposure = 3)

# Create high temperature treatment version
high_temp_rows <- hour3_rows %>%
  mutate(temperature = "42C", time_exposure = 0)

# Combine with original data
segment_data <- full_data %>%
  filter(!timepoint=="3")%>%
  bind_rows(ambient_rows, high_temp_rows) %>%
  arrange(well, as.numeric(timepoint))%>%
  select(!timepoint)
```

Normalize high temperature data to the initial time point. 
```{r}
ambient_rows<-segment_data%>%
  filter(temperature=="18C")

high_rows<-segment_data%>%
  filter(temperature=="42C")%>%
  arrange(date, sample)%>%
  mutate(value=value-first(value))

segment_data<-rbind(ambient_rows, high_rows)
```

Plot separate segments. 

```{r}
#segment_data%>%
#  ggplot(aes(x=time_exposure, y=value, colour=temperature))+
#  facet_wrap(~date*sample)+
#  geom_point()+
#  geom_line(aes(group=temperature))+
#  scale_colour_manual(values=c("darkblue", "darkred"))+
#  theme_classic()
```

```{r}
segment_data%>%
  ggplot(aes(x=time_exposure, y=value, colour=temperature, group=interaction(date,well, temperature)))+
  facet_wrap(~family)+
  geom_point()+
  geom_line()+
  scale_colour_manual(values=c("darkblue", "darkred"))+
  theme_classic()
```

Run a model on metabolic rates. 

```{r}
model<-segment_data%>%
  lmer(value ~ family * temperature + (1|date) + (1|sample:date), data=.)

anova(model)
qqPlot(residuals(model))
```
Significant family x temperature effect on fluorescence values.  

Need to re assess residuals.  

Some oysters have similar slopes while others have a change in slope depending on temperature. 
Calculate slope for each oyster during each temperature phase. 

```{r}
str(segment_data)
segment_data<-as.data.frame(segment_data)
str(segment_data)

# nest data 
nested <- segment_data %>%
  group_by(date, sample, family, temperature) %>%
  nest()

str(nested)
nested$data[[1]]

# group by oyster_id and phase, then fit a linear model of metabolic_rate ~ time
slopes <- nested %>%
  mutate(
    model = map(data, ~ lm(value ~ time_exposure, data = .x)),        # fit model
    slope = map_dbl(model, ~ coef(.x)[["time_exposure"]])                      # extract slope
  ) %>%
  select(date, sample, family, temperature, slope)

# calculate change
slope_summary <- slopes %>%
  pivot_wider(names_from = temperature, values_from = slope) %>%
  mutate(
    absolute_change = `42C` - `18C`,
    percent_change = 100 * (`42C` - `18C`) / `18C`
  )
```

Plot linear model for each sample. 
```{r}
#ggplot(segment_data, aes(x = time_exposure, y = value, color=temperature)) +
#  geom_point() +
#  geom_smooth(aes(group=temperature, color=temperature), method = "lm", se = FALSE) +
#  facet_wrap(~ sample*date, scales = "free_y") +
#  scale_colour_manual(values=c("darkblue", "darkred"))+
#  theme_classic()
```

Plot slope by temperature for each family. 
```{r}
ggplot(slopes, aes(x = temperature, y = slope, color=family)) +
  geom_point() +
  facet_wrap(~family*date)+
  geom_line(aes(group=sample, color=family)) +
  scale_colour_manual(values=c("A"="cyan4", "B"="purple2", "C"="pink", "D"="orange", "E"="darkred"))+
  theme_classic()
```

```{r}
plot2<-ggplot(slopes, aes(x = temperature, y = slope, color=family)) +
  geom_point() +
  facet_wrap(~family)+
  geom_line(aes(group=interaction(sample, date), color=family)) +
  scale_colour_manual(values=c("A"="cyan4", "B"="purple2", "C"="pink", "D"="orange", "E"="darkred"))+
  theme_classic(); plot2

ggsave(plot2, filename="figures/usda-families/fl_family_slopes.png", width=8, height=8, units="in")
```

Plot the percent change in slope for each family from ambient to high temperature. 
```{r}
ggplot(slope_summary, aes(x = family, y = percent_change, color=family)) +
  geom_point() +
  geom_text(aes(label = paste(date, sample, sep = "\n")), 
            vjust = 0.5, size = 3, show.legend = FALSE) +
  geom_hline(yintercept=0, linetype="dashed")+
  scale_colour_manual(values=c("A"="cyan4", "B"="purple2", "C"="pink", "D"="orange", "E"="darkred"))+
  theme_classic()
```

Plot again without labels.  

```{r}
plot3<-ggplot(slope_summary, aes(x = family, y = percent_change, color=family)) +
  geom_point() +
  geom_hline(yintercept=0, linetype="dashed")+
  scale_colour_manual(values=c("A"="cyan4", "B"="purple2", "C"="pink", "D"="orange", "E"="darkred"))+
  theme_classic();plot3

ggsave(plot3, filename="figures/usda-families/fl_family_percent_change.png", width=6, height=6, units="in")
```

Run a model to evaluate the significance of changes in slope between temperatures. 
```{r}
#change to mixed model when we have multiple dates 

model1<-slopes%>%
  lmer(slope ~ family * temperature + (1|date) + (1|sample:date), data=.)

summary(model1)
anova(model1)
qqPlot(residuals(model1))

emm<-emmeans(model1, ~temperature|family)
pairs(emm)

emm<-emmeans(model1, ~family)
pairs(emm)

emm<-emmeans(model1, ~family|temperature)
pairs(emm)
```

Run a model to examine percent change in slope between ambient and high temperature for each family.

```{r}
#change to mixed model when we have multiple dates 

model2<-slope_summary%>%
  lmer(percent_change ~ family + (1|date), data=.)

summary(model2)
anova(model2)
qqPlot(residuals(model2))
```

View posthoc effects. 

```{r}
emm<-emmeans(model2, ~family)
pairs(emm)
```

## Examine mortality effects 

Read in mortality data (0 indicates alive, 1 indicates dead at end of trial). 

```{r}
mortality<-read_xlsx(path="data/usda-families/mortality/mortality.xlsx")%>%
  mutate(date=as.character(date))
```

Join with data frame. 

```{r}
slopes<-left_join(slopes, mortality)
slopes$mortality<-as.character(slopes$mortality)
slope_summary<-left_join(slope_summary, mortality)
slope_summary$mortality<-as.character(slope_summary$mortality)
```

Plot change in metabolic rates colored by whether the oyster died or lived. 

```{r}
plot2a<-ggplot(slopes, aes(x = temperature, y = slope, color=mortality)) +
  geom_point() +
  facet_wrap(~family)+
  geom_line(aes(group=interaction(sample, date), color=mortality)) +
  scale_colour_manual(values=c("0"="black", "1"="red"))+
  theme_classic(); plot2a

ggsave(plot2a, filename="figures/usda-families/fl_family_slopes_mortality.png", width=8, height=8, units="in")
```

```{r}
plot3a<-ggplot(slope_summary, aes(x = family, y = percent_change, color=mortality)) +
  geom_point() +
  geom_hline(yintercept=0, linetype="dashed")+
  scale_colour_manual(values=c("0"="black", "1"="red"))+
  theme_classic();plot3a

ggsave(plot3a, filename="figures/usda-families/fl_family_percent_change_mortality.png", width=6, height=6, units="in")
```

# Absorbance 

## Load data 

```{r}
# Set the folder path
folder_path <- "data/usda-families/absorbance"  

# List all txt files in the folder
file_list <- list.files(path = folder_path, pattern = "\\.txt$", full.names = TRUE, recursive=TRUE)

# Check if any files are found
if (length(file_list) == 0) {
  stop("No txt files found in the specified folder. Check the folder path or file extension.")
}

# Initialize an empty list to store processed data
data_list <- list()

# Loop through each file and load the data
for (file in file_list) {
  # Ensure file is a character string
  if (!is.character(file)) next
  
  # Extract file name without extension
  file_name <- file_path_sans_ext(basename(file))
  
  # Read the text file, skipping the first row (metadata)
  data <- read.delim(file, header = FALSE, skip = 40)
  
  # Assign column names: first column is row labels (A–H), others are 1–12
  colnames(data) <- c("Row", as.character(1:13))
  
  #remove last column
  data<-data%>%select(-ncol(data))
  
  # Convert to long format
  data_long <- data %>%
    pivot_longer(cols = -Row, names_to = "Column", values_to = "Value") %>%
    mutate(
      Column = sprintf("%02d", as.integer(Column)),  # Ensure two-digit column numbers
      Well_ID = paste0(Row, Column),                # Format well ID as "A01", "B02", etc.
      FileName = file_name,                          # Add file name
      date = str_extract(file_name, "^\\d{8}"),     # Extract 8-digit date
      timepoint = str_extract(file_name, "T\\d+") %>% 
        str_remove("T") %>% 
        as.numeric()                                # Convert timepoint to numeric
    ) %>%
    select(FileName, Well_ID, Value, date, timepoint)  # Select relevant columns
  
  # Store the processed data in the list
  data_list[[file_name]] <- data_long
}

# Print an example of processed data
head(data_list[[file_name]])


# Combine all data frames into a single data frame (optional)
combined_data <- bind_rows(data_list, .id = "Source")

# Print the first few rows of the combined data (optional)
head(combined_data)

# Rename columns
combined_data<-combined_data%>%
  rename("well"=Well_ID, resazurin_counts=`Value`)%>%
  mutate(timepoint=as.character(timepoint))

head(combined_data)
```

Load in metadata. 

```{r}
metadata<-read_xlsx(path="data/usda-families/metadata/sample_metadata.xlsx")%>%
  mutate(date=as.character(date))

metadata_temp<-read_xlsx(path="data/usda-families/metadata/temperature_metadata.xlsx")%>%
  mutate(date=as.character(date))%>%mutate(timepoint=as.character(timepoint))
```

Join with data frame and remove any wells that did not have samples. 

```{r}
str(combined_data)
str(metadata)
str(metadata_temp)

full_data<-left_join(combined_data, metadata, by=c("date", "well"))%>%
  filter(!is.na(type))

full_data<-left_join(full_data, metadata_temp, by=c("date", "timepoint"))%>%
  filter(!is.na(type))

head(full_data)
```

Load in size data. 

```{r}
size<-read_xlsx("data/usda-families/size/size.xlsx")%>%
  mutate(date=as.character(date))
```

Join with data. 

```{r}
str(full_data)
str(size)

full_data<-left_join(full_data, size, by=c("date", "sample"))
```

## Prep the data 

Remove sample that was an empty shell. 

```{r}
full_data<-full_data%>%
  filter(!c(date=="20250527" & sample=="B5"))%>%
  filter(!c(date=="20250603" & sample=="B1"))%>%
  filter(!c(date=="20250603" & sample=="C3"))
```

Plot the raw data. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=resazurin_counts, colour=temperature, group=interaction(date, well)))+
  facet_wrap(~family*date)+
  geom_point()+
  geom_line()+
  scale_colour_manual(values=c("darkblue", "darkred"))+
  theme_classic()
```

Calculate size normalized absorbance at each time point normalized to the starting value at time 0. 
```{r}
full_data<-full_data%>%
  group_by(date, well, family)%>%
  arrange(date, well)%>%
  mutate(abs.norm=resazurin_counts/first(resazurin_counts))
```

Plot again. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=abs.norm, colour=temperature, group=interaction(date, well)))+
  facet_wrap(~family*date)+
  geom_point()+
  geom_line()+
  scale_colour_manual(values=c("darkblue", "darkred"))+
  theme_classic()
```

View blanks

```{r}
full_data%>%
  filter(type=="blank")%>%
  ggplot(aes(x=timepoint, y=abs.norm, colour=temperature, group=interaction(date, well)))+
  facet_wrap(~date)+
  geom_point()+
  scale_colour_manual(values=c("darkblue", "darkred"))+
  geom_line()+
  theme_classic()
```

Calculate mean change in blank at each time point. 

```{r}
blanks<-full_data%>%
  filter(type=="blank")%>%
  group_by(date, timepoint)%>%
  summarise(mean_blank=mean(abs.norm));blanks
```

View summarized blank data. 

```{r}
blanks%>%
  ggplot(aes(x=timepoint, y=mean_blank))+
  facet_wrap(~date)+
  geom_point()+
  theme_classic()
```

Subtract blank values from fluorescence values for oysters. 

```{r}
full_data<-left_join(full_data, blanks)

full_data<-full_data%>%
  filter(!type=="blank")%>%
  mutate(abs.corr=abs.norm-mean_blank)
```

Plot again. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=abs.corr, colour=temperature, group=interaction(date, well)))+
  facet_wrap(~family*date)+
  geom_point()+
  geom_line()+
  scale_colour_manual(values=c("darkblue", "darkred"))+
  theme_classic()
```

Size normalize data. I am going to use area first. 
```{r}
#normalize by area
full_data<-full_data%>%
  mutate(abs.corr.mm2=abs.corr/area)

#normalize by length=
full_data<-full_data%>%
  mutate(abs.corr.mm=abs.corr/length.mm)
```

Plot again. 

Plot by area normalized. 
```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=abs.corr.mm2, colour=temperature, group=interaction(date, well)))+
  facet_wrap(~family*date)+
  geom_point()+
  geom_line()+
  scale_colour_manual(values=c("darkblue", "darkred"))+
  theme_classic()
```

Plot by length
```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=abs.corr.mm, colour=temperature, group=interaction(date, well)))+
  facet_wrap(~family*date)+
  geom_point()+
  geom_line()+
  scale_colour_manual(values=c("darkblue", "darkred"))+
  theme_classic()
```

We will use area moving forward. Both seem to give us very similar answers and area will be more representative of total size.  

Plot again. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=abs.corr.mm2, colour=temperature, group=interaction(date, well)))+
  facet_wrap(~family*date)+
  geom_point()+
  geom_line()+
  scale_colour_manual(values=c("darkblue", "darkred"))+
  theme_classic()
```

Remove unnecessary columns. 

```{r}
str(full_data)

full_data<-full_data%>%
  select(!Source)%>%
  select(!FileName)%>%
  select(!resazurin_counts)%>%
  select(!type)%>%
  select(!mean_blank)%>%
  select(!area)%>%
  select(!length.mm)%>%
  select(!abs.corr)%>%
  select(!abs.norm)

str(full_data)

full_data<-full_data%>%
  rename(value=abs.corr.mm2)
```


## Models and Plots    

Plot raw data. 
```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=value, colour=family, group=interaction(date, family, well)))+
  facet_wrap(~date)+
  geom_point()+
  geom_vline(xintercept=4, linetype="dashed")+
  scale_colour_manual(values=c("A"="cyan4", "B"="purple2", "C"="pink", "D"="orange", "E"="darkred"))+
  geom_line()+
  theme_classic()
```

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=value, colour=temperature, group=interaction(date, well)))+
  facet_wrap(~family)+
  geom_point()+
  geom_vline(xintercept=4, linetype="dashed")+
  scale_colour_manual(values=c("darkblue", "darkred"))+
  geom_line()+
  theme_classic()
```

## Calculate change due to temperature exposure   

Add a column for time of exposure. 

```{r}
full_data<-full_data%>%
  mutate(time_exposure=case_when(
    timepoint=="0" ~ 0,
    timepoint=="1" ~ 1,
    timepoint=="2" ~ 2,
    timepoint=="3" ~ 3,
    timepoint=="4" ~ 1,
    timepoint=="5" ~ 2
  ))

#replicate data from time point 3 as time of exposure 0 for 42C treatment and time of exposure 3 for 18C treatment

str(full_data)

# Create a copy of the hour 3 rows
hour3_rows <- full_data %>% 
  filter(timepoint == "3")

# Create ambient treatment version
ambient_rows <- hour3_rows %>%
  mutate(temperature = "18C", timepoint = "3", time_exposure = 3)

# Create high temperature treatment version
high_temp_rows <- hour3_rows %>%
  mutate(temperature = "42C", time_exposure = 0)

# Combine with original data
segment_data <- full_data %>%
  filter(!timepoint=="3")%>%
  bind_rows(ambient_rows, high_temp_rows) %>%
  arrange(well, as.numeric(timepoint))%>%
  select(!timepoint)
```

Normalize high temperature data to the initial time point. 
```{r}
ambient_rows<-segment_data%>%
  filter(temperature=="18C")

high_rows<-segment_data%>%
  filter(temperature=="42C")%>%
  arrange(date, sample)%>%
  mutate(value=value-first(value))

segment_data<-rbind(ambient_rows, high_rows)
```

Plot separate segments. 

```{r}
segment_data%>%
  ggplot(aes(x=time_exposure, y=value, colour=temperature))+
  facet_wrap(~date*sample)+
  geom_point()+
  geom_line(aes(group=temperature))+
  scale_colour_manual(values=c("darkblue", "darkred"))+
  theme_classic()
```

Some oysters have similar slopes while others have a change in slope depending on temperature. 
Calculate slope for each oyster during each temperature phase. 

```{r}
str(segment_data)
segment_data<-as.data.frame(segment_data)
str(segment_data)

# nest data 
nested <- segment_data %>%
  group_by(date, sample, family, temperature) %>%
  nest()

str(nested)
nested$data[[1]]

# group by oyster_id and phase, then fit a linear model of metabolic_rate ~ time
slopes <- nested %>%
  mutate(
    model = map(data, ~ lm(value ~ time_exposure, data = .x)),        # fit model
    slope = map_dbl(model, ~ coef(.x)[["time_exposure"]])                      # extract slope
  ) %>%
  select(date, sample, family, temperature, slope)

# calculate change
slope_summary <- slopes %>%
  pivot_wider(names_from = temperature, values_from = slope) %>%
  mutate(
    absolute_change = `42C` - `18C`,
    percent_change = 100 * (`42C` - `18C`) / `18C`
  )
```

Plot linear model for each sample. 
```{r}
ggplot(segment_data, aes(x = time_exposure, y = value, color=temperature)) +
  geom_point() +
  geom_smooth(aes(group=temperature, color=temperature), method = "lm", se = FALSE) +
  facet_wrap(~ sample*date, scales = "free_y") +
  scale_colour_manual(values=c("darkblue", "darkred"))+
  theme_classic()
```

Plot slope by temperature for each family. 
```{r}
ggplot(slopes, aes(x = temperature, y = slope, color=family)) +
  geom_point() +
  facet_wrap(~family*date)+
  geom_line(aes(group=sample, color=family)) +
  scale_colour_manual(values=c("A"="cyan4", "B"="purple2", "C"="pink", "D"="orange", "E"="darkred"))+
  theme_classic()

plot2a<-ggplot(slopes, aes(x = temperature, y = slope, color=family)) +
  geom_point() +
  facet_wrap(~family)+
  geom_line(aes(group=interaction(date, sample), color=family)) +
  scale_colour_manual(values=c("A"="cyan4", "B"="purple2", "C"="pink", "D"="orange", "E"="darkred"))+
  theme_classic(); plot2a

ggsave(plot2a, filename="figures/usda-families/ab_family_slopes.png", width=8, height=8, units="in")
```

Plot the percent change in slope for each family from ambient to high temperature. 
```{r}
plot3a<-ggplot(slope_summary, aes(x = family, y = percent_change, color=family)) +
  geom_point() +
  geom_hline(yintercept=0, linetype="dashed")+
  scale_colour_manual(values=c("A"="cyan4", "B"="purple2", "C"="pink", "D"="orange", "E"="darkred"))+
  theme_classic()+
  ylim(-900,2000);plot3a #removing an outlier 

ggsave(plot3a, filename="figures/usda-families/ab_family_percent_change.png", width=8, height=8, units="in")
```

Run a model to evaluate the significance of changes in slope between temperatures. 
```{r}
#change to mixed model when we have multiple dates 

model1<-slopes%>%
  lm(slope ~ family * temperature, data=.)

summary(model1)
anova(model1)
qqPlot(model1$residuals)

emm<-emmeans(model1, ~temperature|family)
pairs(emm)
```

Run a model to examine percent change in slope between ambient and high temperature for each family. 

```{r}
#change to mixed model when we have multiple dates 

model2<-slope_summary%>%
  lm(percent_change ~ family, data=.)

summary(model2)
anova(model2)
qqPlot(model2$residuals)
```

Residuals meet assumptions.  

View posthoc effects. 

```{r}
emm<-emmeans(model2, ~family)
pairs(emm)
```

## Examine mortality effects 

Read in mortality data (0 indicates alive, 1 indicates dead at end of trial). 

```{r}
mortality<-read_xlsx(path="data/usda-families/mortality/mortality.xlsx")%>%
  mutate(date=as.character(date))
```

Join with data frame. 

```{r}
slopes<-left_join(slopes, mortality)
slopes$mortality<-as.character(slopes$mortality)
slope_summary<-left_join(slope_summary, mortality)
slope_summary$mortality<-as.character(slope_summary$mortality)
```

Plot change in metabolic rates colored by whether the oyster died or lived. 

```{r}
plot2a<-ggplot(slopes, aes(x = temperature, y = slope, color=mortality)) +
  geom_point() +
  facet_wrap(~family)+
  geom_line(aes(group=interaction(sample, date), color=mortality)) +
  scale_colour_manual(values=c("0"="black", "1"="red"))+
  theme_classic(); plot2a

ggsave(plot2a, filename="figures/usda-families/ab_family_slopes_mortality.png", width=8, height=8, units="in")
```

```{r}
plot3a<-ggplot(slope_summary, aes(x = family, y = percent_change, color=mortality)) +
  geom_point() +
  geom_hline(yintercept=0, linetype="dashed")+
  scale_colour_manual(values=c("0"="black", "1"="red"))+
  theme_classic()+
  ylim(-3000,3000);plot3a

ggsave(plot3a, filename="figures/usda-families/ab_family_percent_change_mortality.png", width=6, height=6, units="in")
```


