---
title: "10K seed resazurin survival analysis"
author: "AS Huffmyer"
date: '2024'
output:
  md_document: default
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 6
    toc_float: true
  github_document: null
editor_options:
  chunk_output_type: console
---

TO DO: Examine if mortality relates to size, then add size data as random/covariate effect for mortality 

# Set up 

Set up workspace, set options, and load required packages.    
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries. 
```{r}
library(tidyverse)
library(ggplot2)
library(survival)
library(readxl)
library(ggsurvfit)
library(gtsummary)
library(cardx)
library(cowplot)
```

# Read data 

Read in data. 
```{r}
data<-read_excel("data/survival/survival_resazurin.xlsx")
```

Turn into long format. 
```{r}
data<-data%>%
  pivot_longer(names_to="time", values_to="status", cols=`0`:`24`)%>%
  mutate(time=as.numeric(time))
```

Create data frames with control and high temperature subsets. 
```{r}
control<-data%>%filter(temperature=="18C")
stress<-data%>%filter(temperature=="42C")
```


# Generate Kaplan Meier survival curves 

## Control temperature 18C

```{r}
s1 <- survfit(Surv(time, status) ~ hardening, data = control)
str(s1)
```

Plot the survival function
```{r}
survfit2(Surv(time, status) ~ hardening+temperature, data = control) %>% 
  ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability"
  )
```

Estimate survival at 24 hours. 
```{r}
summary(survfit(Surv(time, status) ~ hardening, data = control), times = 24)
```

Estimated survival is ~87% in control temperature. 

Use a log rank model to determine statistical differences in curves. 
```{r}
survdiff(Surv(time, status) ~ hardening, data = control)

# Call:
# survdiff(formula = Surv(time, status) ~ hardening, data = control)
# 
#                                     N Observed Expected (O-E)^2/E (O-E)^2/V
# hardening=control                 360       18     19.1   0.06234   0.11188
# hardening=fresh-water             240       12     12.7   0.04156   0.06630
# hardening=fresh-water-temperature 240       16     12.7   0.84156   1.34252
# hardening=immune                  240       13     12.7   0.00584   0.00932
# hardening=temperature             240       11     12.7   0.23442   0.37396
# 
#  Chisq= 1.5  on 4 degrees of freedom, p= 0.8 
```
There is no difference in survival at 18C at p=0.8. 

Analyze again with a Cox proportional hazards model. 
```{r}
coxph(Surv(time, status) ~ hardening, data = control)

# coxph(formula = Surv(time, status) ~ hardening, data = control)
# 
#                                      coef exp(coef) se(coef)      z     p
# hardeningfresh-water              0.02005   1.02025  0.37269  0.054 0.957
# hardeningfresh-water-temperature  0.36099   1.43475  0.34368  1.050 0.294
# hardeningimmune                   0.07687   1.07991  0.36398  0.211 0.833
# hardeningtemperature             -0.06777   0.93447  0.38271 -0.177 0.859
# 
# Likelihood ratio test=1.57  on 4 df, p=0.8147
# n= 1320, number of events= 70  

coxph(Surv(time, status) ~ hardening, data = control) %>% 
  tbl_regression(exp = TRUE) 
```
Control is the reference level. No differences in survival at 18C. 

# Generate Kaplan Meier survival curves 

## Stress temperature 42C

```{r}
s2 <- survfit(Surv(time, status) ~ hardening, data = stress)
str(s2)
```

Plot the survival function
```{r}
survfit2(Surv(time, status) ~ hardening+temperature, data = stress) %>% 
  ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability"
  )
```

Estimate survival at 24 hours. 
```{r}
summary(survfit(Surv(time, status) ~ hardening, data = stress), times = 24)
```

Estimated stress survival is 12-26%. Temperature hardened is the lowest.    

Use a log rank model to determine statistical differences in curves. 
```{r}
survdiff(Surv(time, status) ~ hardening, data = stress)

# Call:
# survdiff(formula = Surv(time, status) ~ hardening, data = stress)
# 
#                                     N Observed Expected (O-E)^2/E (O-E)^2/V
# hardening=control                 360      116    133.1     2.195     5.048
# hardening=fresh-water             240       83     88.7     0.370     0.756
# hardening=fresh-water-temperature 240       83     88.7     0.370     0.756
# hardening=immune                  240       97     88.7     0.771     1.577
# hardening=temperature             240      109     88.7     4.632     9.470
# 
#  Chisq= 13.9  on 4 degrees of freedom, p= 0.007 
```
There is a significant difference in survival at p=0.007. 

Analyze again with a Cox proportional hazards model. 
```{r}
coxph(Surv(time, status) ~ hardening, data = stress)

# Call:
# coxph(formula = Surv(time, status) ~ hardening, data = stress)
# 
#                                     coef exp(coef) se(coef)     z       p
# hardeningfresh-water             0.08431   1.08797  0.14377 0.586 0.55759
# hardeningfresh-water-temperature 0.04540   1.04644  0.14380 0.316 0.75224
# hardeningimmune                  0.20959   1.23317  0.13763 1.523 0.12780
# hardeningtemperature             0.40947   1.50602  0.13343 3.069 0.00215
# 
# Likelihood ratio test=11.11  on 4 df, p=0.02538
# n= 1320, number of events= 488 

coxph(Surv(time, status) ~ hardening, data = stress) %>% 
  tbl_regression(exp = TRUE) 
```
Control is the reference level. There is a significant difference between control and temperature hardening survival. 

# Generate plots 
 
Set theme. 
```{r}
my_theme<-theme_classic()
```

Control
```{r}
plot1<-survfit2(Surv(time, status) ~ hardening, data = control) %>% 
  ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title="18°C"
  )+
  ylim(0,1)+
  my_theme+
  geom_text(x=10, y=0.2, label="Cox PH p=0.8")+
  theme(legend.position="none")
```

Stress
```{r}
plot2<-survfit2(Surv(time, status) ~ hardening, data = stress) %>% 
  ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title = "42°C"
  )+
  ylim(0,1)+
  my_theme+
  geom_text(x=10, y=0.2, label="Cox PH p=0.025")+
  geom_text(x=10, y=0.15, label="Temperature vs control p=0.002")+
  geom_text(x=10, y=0.10, label="All others vs control p>0.05")+
  theme(legend.position="right")
```

Assemble plot
```{r}
plots<-plot_grid(plot1, plot2, rel_widths=c(0.65,1), ncol=2)

ggsave(plots, filename="figures/survival/KMcurves.png", width=10, height=4)
```


