---
title: Open vs closed oyster testing analysis
author: "AS Huffmyer"
date: '2025'
output:
  github_document: null
  md_document: default
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 6
    toc_float: true
editor_options: 
  chunk_output_type: console
---

# Set up 

Set up workspace, set options, and load required packages.    
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries. 
```{r}
library(MASS) 
library(tidyverse)
library(ggplot2)
library(readxl)
library(cowplot)
library(lme4)
library(lmerTest)
library(car)
library(effects)
library(emmeans)
library(RColorBrewer)
library(pROC)
```

# Load data 

Read in testing data
```{r}
#read in files
data <- read_excel("data/testing/open_closed_trials.xlsx")
```

# Data preparation 

Plot the data. 

```{r}
data%>%
  ggplot(aes(x=timepoint, y=value, colour=treatment, group=interaction(trial,sample)))+
  geom_point()+
  geom_line()+
  theme_classic()
```

Calculate fluorescence at each time point normalized to the starting value at time 0. 

```{r}
data<-data%>%
  group_by(trial, sample, treatment, status, final.status)%>%
  arrange(trial, sample)%>%
  mutate(fluorescence.norm=value/first(value))
```

Plot again 

```{r}
data%>%
  ggplot(aes(x=timepoint, y=fluorescence.norm, colour=treatment, group=interaction(trial,sample)))+
  geom_point()+
  geom_line()+
  theme_classic()
```

Plot within temperature between open and closed oysters. 

```{r}
data%>%
  ggplot(aes(x=timepoint, y=fluorescence.norm, colour=final.status, group=interaction(trial,sample)))+
  facet_wrap(~treatment)+
  geom_point()+
  geom_line()+
  theme_classic()
```

Calculate total fluorescence for each sample as the value at the last time point.  

```{r}
data<-data%>%
  group_by(trial, sample, treatment, final.status)%>%
  arrange(trial, sample)%>%
  mutate(fluorescence.total=last(fluorescence.norm))%>%
  select(!fluorescence.norm)%>%
  select(!timepoint)%>%
  select(!status)%>%
  select(!value)%>%
  unique()
```

Plot total fluorescence by temperature and status. 

```{r}
data%>%
  filter(fluorescence.total<9)%>% #remove outlier
  ggplot(aes(x=final.status, y=fluorescence.total, group=interaction(trial,sample)))+
  facet_wrap(~treatment)+
  geom_point()+
  theme_classic()+
  ylab("Change in normalized fluorescence")+
  xlab("Behavior")+
  theme(legend.position="none")
```

Run a statistical test. 

```{r}
model<-aov(fluorescence.total ~ treatment * final.status, data = data)
summary(model)
```

There is an effect of open vs closed that is consistent between treatments. A higher metabolic rate is not indicative of higher metabolism due to being open.  
